---
title: "PPT-ANPP"
author: "OLH"
date: "2026-04"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, supportR, lubridate)

`%!in%` = Negate(`%in%`)

```

# Create Folders on Desktop
```{r}
dir.create(file.path("deluge"), showWarnings = F)
dir.create(file.path("deluge", 'data'), showWarnings = F)
dir.create(file.path("deluge", 'precip_data'), showWarnings = F)
dir.create(file.path("deluge", 'data', 'R_exports'), showWarnings = F)
```

# Download ANPP, PPT, and Treatment matching table files
```{r}
##### Deluge characteristics
trt.data <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3")) %>%
  dplyr::filter(name == "deluge_characteristics_mergeable.csv")

# Did that work?
trt.data 

# Download the excluded treatmetn file
googledrive::drive_download(file = trt.data$id, overwrite = T, type = "csv",
                            path = file.path("deluge","data","R_exports", trt.data$name))

# Read in hoover ppt
trt.data <- read.csv(file = file.path("deluge","data","R_exports", "deluge_characteristics_mergeable.csv"))


##### Precip
ppt <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/18toIfC8jV7s8ZzAD0SV9CWL4SvGgn-RO")) %>%
  dplyr::filter(name == "precip_data_cleaned_trts.csv")

# Did that work?
ppt 

# Download the excluded treatmetn file
googledrive::drive_download(file = ppt$id, overwrite = T, type = "csv",
                            path = file.path("deluge","precip_data", ppt$name))

# Read in hoover ppt
ppt <- read.csv(file = file.path("deluge","precip_data", "precip_data_cleaned_trts.csv"))

##### ANPP
anpp <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3")) %>%
  dplyr::filter(name == "ANPP_filtered.csv")

# Did that work?
anpp 

# Download the excluded treatmetn file
googledrive::drive_download(file = anpp$id, overwrite = T, type = "csv",
                            path = file.path("deluge","data","R_exports", anpp$name))

# Read in hoover ppt
anpp <- read.csv(file = file.path("deluge","data","R_exports", "ANPP_filtered.csv"))


# TREATMENT NAME UPDaTE
names <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1fpoG0wLLqjFLkucR2ZlNMon31DIW1GH-")) %>%
  dplyr::filter(name == "treatment_codes_byData")

# Did that work?
names 

# Download the excluded treatmetn file
googledrive::drive_download(file = names$id, overwrite = T, type = "csv",
                            path = file.path("deluge", names$name))

# Read in hoover ppt
names.update <- read.csv(file = file.path("deluge","treatment_codes_byData.csv"))

# read in differences from long ta
lta <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1fpoG0wLLqjFLkucR2ZlNMon31DIW1GH-")) %>%
  dplyr::filter(name == "treatmetn_LTA_Change.csv")

googledrive::drive_download(file = lta$id, overwrite = T, type = "csv",
                            path = file.path("deluge", lta$name))

# Read in hoover ppt
trt.data <- read.csv(file = file.path("deluge", "treatmetn_LTA_Change.csv"))

```

# prelim ANPP data viz
```{r}
str(anpp)
anpp.tot <- anpp %>%
  filter(funct_group == "ANPP")
ggplot(anpp.tot, aes(year, ANPP, color = Study_shorthand))+
  geom_jitter(alpha = 0.6)+
  theme_bw()+
  ylim(0, 300)

```


# Clean and merge data

```{r}
# 1) fix/update treatment names on the precip data
names.update  <- names.update %>%
  filter(raw_filename == "precip_data_cleaned_trts.csv") %>%
  rename(Study = "Full.Study", 
         Treatment_raw = "trt_names") %>%
  select(c(Study, Study_shorthand, Our_trt, Treatment_raw)) %>%
  mutate(across(where(is.character), ~gsub("\u00A0", " ", .x)))

ppt.2 <- ppt %>%
  mutate(across(c(Study, Treatment_raw), str_trim)) %>%
  left_join(
    names.update, #%>% mutate(across(c(Study, Treatment_raw), str_trim)),
    by = c("Study", "Treatment_raw")
  )

## Merge with ANPP and deluge data
# str(anpp)
# str(ppt.2)
# str(trt.data)

trt <- trt.data %>%
  # 1. Remove rows where every cell is NA OR a blank string ""
  filter(!if_all(everything(), ~ is.na(.x) | .x == "")) %>%
  
  # 2. Remove columns where every cell is NA OR a blank string ""
  select(where(~ !all(is.na(.x) | .x == "")))


ppt.3 <- ppt.2 %>%
  select(-c(Study, Treatment_raw)) %>%
  rename(trt = "Our_trt")

anpp.2 <- anpp %>%
  dplyr::filter(!year %in% c("July-Sept Regrowth 2021","July 2021")) %>%
  mutate(year = as.integer(ifelse(year == "Sept 2021", "2021", year))) %>%
  filter(funct_group == "ANPP") %>%
  group_by(Study_shorthand, year, trt) %>%
  summarize(mean_ANPP = mean(ANPP), se_ANPP = (sd(ANPP))/(sqrt(n())))

## JOIN them all
anpp.ppt <- anpp.2%>%
  left_join(trt) %>%
  left_join(ppt.3) %>%
  mutate(ann_ppt_13 = ann_ppt + m9_prev, 
         ann_ppt_13 = ifelse(year == 2022 & trt == "CON_DEL" & Study_shorthand == "Linabury_ChANGE", ann_ppt_13 + 42, ann_ppt_13))
```

## Start looking at some visuals
```{r}
ggplot(anpp.ppt, aes(ann_ppt_13, mean_ANPP))+
  geom_point(size=3, aes(color = Study_shorthand)) +
  theme_bw() +
  geom_smooth(method="lm") 
library(car)
model <- lm(mean_ANPP ~ ann_ppt_13, data = anpp.ppt)
summary(model)


ggplot(anpp.ppt, aes(gs_ppt, mean_ANPP))+
  geom_point(size=3, aes(color = Study_shorthand)) +
  theme_bw() +
  geom_smooth(method="lm") 

ggplot(anpp.ppt, aes(ann_ppt_prev, mean_ANPP))+
  geom_point(size=3, aes(color = Study_shorthand)) +
  theme_bw() +
  geom_smooth(method="lm") 

```

# RUE
```{r}
rue <- anpp.ppt %>%
  mutate(rue = mean_ANPP/ann_ppt_13, 
         del = as.numeric(ifelse(trt %in% c("CON", "DR"), "0", "1")))

ggplot(rue, aes(ann_ppt_13, rue))+
  geom_point(size=3, aes(color = Study_shorthand)) +
  theme_bw() +
  geom_smooth(method="lm") +
  facet_wrap(~Deluge.Timing)

ggplot(rue, aes(mean_ANPP, rue))+
  geom_point(size=3, aes(color = Study_shorthand)) +
  theme_bw() +
  geom_smooth(method="lm")

```

# looking at LRR
```{r}
con <- rue %>%
  filter(del == 0) %>%
  mutate(pairing = str_sub(trt, 1, 2)) %>%
  mutate(spring_ppt = m4 + m5 + m6) %>%
  select(pairing,ann_ppt_13, Study_shorthand, year, spring_ppt, rue, mean_ANPP, se_ANPP) %>%
  rename(ctrl_anpp = "mean_ANPP", 
         ctrl_se = "se_ANPP", 
         ctrl_rue = "rue", 
         ctrl_ppt = "ann_ppt_13")

trts <-rue %>%
  filter(del==1) %>%
  mutate(pairing = str_sub(trt, 1, 2), 
         pairing = ifelse(pairing == "DR", "DR", "CO")) %>%
  select(pairing,ann_ppt_13, Study_shorthand, year, rue, mean_ANPP, se_ANPP) %>%
  rename(del_anpp = "mean_ANPP", 
         del_se = "se_ANPP", 
         del_rue = "rue", 
         del_ppt = "ann_ppt_13")

paired <- left_join(con, trts)

paired.2 <- paired %>%
  mutate(lrr_anpp = log(del_anpp/ctrl_anpp), 
         lrr_rue = log(del_rue/ctrl_rue), 
         del_size = del_ppt - ctrl_ppt)

ggplot(paired.2, aes(lrr_rue))+
  geom_histogram()

ggplot(paired.2, aes(del_size, lrr_anpp, color = Study_shorthand))+
  geom_point()

ggplot(rue, aes(ann_ppt_13, mean_ANPP, color = Study_shorthand))+
  geom_point()+
  theme_bw() +
  facet_wrap(~del)
  
model <- lm(data =rue, mean_ANPP ~ ann_ppt_13 + as.factor(del)) 
summary(model)
```

# Going to do a meta-regression looking at the LRR ~ deluge_size + deviation of spring precip (Apr-May-June) from LTA + timing + (1|year/location)

```{r}
# 1 read in the long-term data precip
lta <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/18toIfC8jV7s8ZzAD0SV9CWL4SvGgn-RO")) %>%
  dplyr::filter(name == "EDGE_MS_full_NOAA_EDGE_precipitation.csv")

# Did that work?
lta

# Download the excluded treatmetn file
googledrive::drive_download(file = lta$id, overwrite = T, type = "csv",
                            path = file.path("deluge","precip_data", lta$name))

# Read in hoover ppt
lta <- read.csv(file = file.path("deluge","precip_data", "EDGE_MS_full_NOAA_EDGE_precipitation.csv"))

lta2 <- lta %>%
  filter(Site == "SGS") %>%
  filter(amb_drought == "amb_adjusted") %>%
  select(DATE, daily.tot) %>%
  mutate(DATE = as.Date(DATE, format = "%Y-%m-%d"), 
         year = year(DATE), 
         month = month(DATE)) %>%
  group_by(year, month) %>%
  #filter(month %in% c(4,5,6)) %>%
  summarize(monthly_ppt = sum(daily.tot)) %>%
  ungroup()%>%
  filter(month %in% c(4,5,6)) %>%
  group_by(month) %>%
  summarize(monthly_average = mean(monthly_ppt)) 

lta_mean_spring <- 36.47203 + 59.60872 + 47.97409

# ggplot(lta2, aes(year, spring_ppt))+
#   geom_point()+geom_line()+
#   theme_bw()
#   

MAP <- sum(lta2$monthly_average, na.rm = TRUE)
```

# get the model together

```{r}

# 2 set up model
library(lmerTest)
library(lme4)
library(car)        # for Type III ANOVA and VIF
library(performance) # convenient assumption checks
library(DHARMa)  

paired.2$lta_spring <- lta_mean

paired.3 <- paired.2 %>%
  mutate(spring_dev = spring_ppt - lta_spring,
         timing = as.factor(timing), 
         location = as.factor(case_when(
           Study_shorthand == "CCE" ~ "CCE", 
           Study_shorthand == "Linabury_ChANGE" ~ "CHANGE",
           TRUE ~ "DEX"
         )), 
         year = as.factor(year)
  ) %>%
  filter(Study_shorthand != "Linabury_ChANGE")


fall <- paired.3 %>%
  filter(timing == "fall")

gs <- paired.3 %>%
  filter(timing != "fall")

ggplot(gs, aes(spring_dev, lrr_anpp, color = Study_shorthand))+
  geom_point(size =2)+
  theme_bw()

ggplot(paired.3, aes(timing, lrr_anpp, color = Study_shorthand))+
  geom_point()+
  theme_bw()

ggplot(paired.3, aes(del_size, lrr_anpp, color = Study_shorthand))+
  geom_point()+
  theme_bw()



model_lrr <- lmer(data = gs, lrr_anpp ~  spring_dev + del_size + (1|location) + (1|year))
summary(model_lrr)


# --- ANOVA on fixed effects ---
# If you loaded lmerTest before fitting, anova() uses Satterthwaite df by default
anova(model_lrr)

# Type III ANOVA (use if you have interactions or want marginal tests)
anova(model_lrr, type = 3, ddf = "Kenward-Roger")

# car::Anova is another option
Anova(model_lrr, type = "III", test.statistic = "F")

# Significance test of the random effects (likelihood ratio test)
ranova(model_lrr)

# --- Assumption checks ---

# Residuals vs fitted (linearity, homoscedasticity)
plot(model_lrr)

# QQ plot of residuals (normality of residuals)
qqnorm(resid(model_lrr)); qqline(resid(model_lrr))

# QQ plots of random effects (normality of random intercepts)
lattice::qqmath(ranef(model_lrr))

# Scale-location plot (homoscedasticity)
plot(model_lrr, sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p", "smooth"))

# Residuals vs each fixed-effect predictor (look for missed nonlinearity)
plot(model_lrr, resid(.) ~ del_size)
plot(model_lrr, resid(.) ~ spring_dev)
plot(model_lrr, resid(., type = "pearson") ~ timing)

# Multicollinearity among fixed effects
vif(model_lrr)

# 3. DHARMa simulation-based residuals (more robust than QQ-eyeballing)
sim_res <- simulateResiduals(model_lrr)
plot(sim_res)
testResiduals(sim_res)   # tests uniformity, dispersion, outliers


```


# Making a summary of the precipitation tables
* Want to know MAP (Water-year? sept-sept? I think water-year except for the fall deluges) and then could have an asterisk
* Question - are Greg's DR treatments just those iwht the drought legacy?
* early season conditions in general correspond well with the year (91.5 R2 with a correlation)

```{r}
# Just going to select the controls - DR or CON
ppt.cons <- ppt.3 %>%
  filter(trt %in% c("CON", "DR")) %>%
  mutate(MAP = MAP, 
         diff_LTA = ann_ppt - MAP, 
         early_season_lta = lta_mean_spring, 
         early_season = m4 + m5 + m6, 
         early_season_diff = early_season - early_season_lta) %>%
  select(Study_shorthand, year, trt, diff_LTA)


ggplot(ppt.cons, aes(diff_LTA, early_season_diff))+
  geom_point()
cor(ppt.cons$diff_LTA, ppt.cons$early_season_diff)


```

