---
title: "ANPP"
author: "OLH"
date: "2026-04"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, supportR, lubridate)

`%!in%` = Negate(`%in%`)

```

# Create Folders on Desktop
```{r}
dir.create(file.path("deluge"), showWarnings = F)
dir.create(file.path("deluge", 'data'), showWarnings = F)
dir.create(file.path("deluge", 'precip_data'), showWarnings = F)
dir.create(file.path("deluge", 'data', 'R_exports'), showWarnings = F)
```

# Download ANPP data
```{r}
##### ANPP
anpp <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3")) %>%
  dplyr::filter(name == "ANPP_filtered.csv")

# Did that work?
anpp 

# Download the excluded treatmetn file
googledrive::drive_download(file = anpp$id, overwrite = T, type = "csv",
                            path = file.path("deluge","data","R_exports", anpp$name))

# Read in hoover ppt
anpp <- read.csv(file = file.path("deluge","data","R_exports", "ANPP_filtered.csv"))

tooley <- anpp %>%
  filter(Study_shorthand == "Tooley_DRE")
```

# Visualize the ANPP by type too
```{r}
unique(anpp$funct_group)
str(anpp)
anpp.2 <- anpp %>%
  # fix the years for consistency of measurement
  filter(year %!in% c("July-Sept Regrowth 2021","July 2021")) %>%
  mutate(year = ifelse(year == "Sept 2021", "2021", year)) %>%
  # create a new func group that's just grass, anpp, or total
  mutate(func = case_when(
    funct_group == "ANNUALS" ~ "GRASS", 
    funct_group == "Annual" ~ "GRASS", 
    funct_group == "Forb" ~ "FORB",
    funct_group == "C3" ~ "GRASS",
    funct_group == "C4" ~ "GRASS",
    funct_group == "C4.ANNUALS" ~ "GRASS",
    TRUE ~ funct_group
  )) %>%
  # summarize into grass, annual, or total
  group_by(Study_shorthand, year, Block, plot, trt, func) %>%
  summarize(anpp = sum(ANPP)) %>%
  # get rid of woody
  filter(func != "WOODY") %>%
  filter(func != "Woody")

ggplot(anpp.2, aes(year, anpp, color = Study_shorthand))+
  geom_jitter(alpha = 0.6)+
  theme_bw()+
  facet_wrap(~func)
```


# read in table to look at comparisons
```{r}
# read in differences from long ta
lta <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1fpoG0wLLqjFLkucR2ZlNMon31DIW1GH-")) %>%
  dplyr::filter(name == "treatmetn_LTA_Change.csv")

googledrive::drive_download(file = lta$id, overwrite = T, type = "csv",
                            path = file.path("deluge", lta$name))

# Read in hoover ppt
trt.data <- read.csv(file = file.path("deluge", "treatmetn_LTA_Change.csv"))

```


# summarize across func groups and determine a log response ratio
```{r}

# create a new dataframe that determines if it is a control or a deluge treatment and then summarizes to the treatment year level, averaging across plots and then blocks where relevant
str(anpp.2)
anpp.3 <- anpp.2%>%
  mutate(del = as.numeric(ifelse(trt %in% c("CON", "DR"), "0", "1"))) %>%
  group_by(Study_shorthand, year, Block, trt, func, del) %>%
  summarize(mean_ANPP = mean(anpp), se_ANPP = (sd(anpp))/(sqrt(n())))

block_summary <- anpp.3%>%
  filter(!is.na(Block)) %>%
  group_by(Study_shorthand, year, trt, func, del) %>%
  summarise(mean_anpp = mean(mean_ANPP), se_ANPP = (sd(mean_ANPP))/(sqrt(n())), .groups = "drop") %>%
  rename(mean_ANPP = "mean_anpp")

anpp.4 <- anpp.3 %>%
  filter(is.na(Block)) %>%
  ungroup()%>%
  select(-Block) %>%
  rbind(block_summary)

con <- anpp.4 %>%
  filter(del == 0) %>%
  mutate(pairing = str_sub(trt, 1, 2)) %>%
  select(pairing, Study_shorthand, year, func, mean_ANPP, se_ANPP) %>%
  rename(ctrl_anpp = "mean_ANPP", 
         ctrl_se = "se_ANPP")

trts <- anpp.4  %>%
  filter(del==1) %>%
  mutate(pairing = str_sub(trt, 1, 2), 
         pairing = ifelse(pairing == "DR", "DR", "CO")) %>%
  select(pairing, trt, Study_shorthand, year, func, mean_ANPP, se_ANPP) %>%
  rename(del_anpp = "mean_ANPP", 
         del_se = "se_ANPP")

paired <- left_join(con, trts)

paired.2 <- paired %>%
  mutate(lrr_anpp = log(del_anpp/ctrl_anpp))

ggplot(paired.2, aes(year, lrr_anpp, color = Study_shorthand)) +
  geom_point()+
  theme_bw()+
  facet_wrap(~func)
```

# look at overall mean log response ratio by func
```{r}
func.rr <- paired.2 %>%
  # filter to not include fall
  #filter(Study_shorthand %in% c("Hajek2024", "Tooley_DRE")) %>%
  group_by(func) %>%
  summarize(mean_lrr = mean(lrr_anpp), se_lrr = (sd(lrr_anpp))/(sqrt(n()))) 

ggplot() +
  geom_jitter(data = paired.2, aes(x = lrr_anpp, y = func, color = Study_shorthand, shape = func), 
           alpha = 0.6, size = 2.2) +
    geom_point(data = func.rr, aes(x = mean_lrr, y = func, shape = func), size = 3) +
  geom_errorbar(data = func.rr, aes(x = mean_lrr, y = func,
                                     xmin = mean_lrr - se_lrr, 
                                     xmax = mean_lrr + se_lrr),
                width = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw(base_size = 14) +
  ylab("Functional Group") +
  xlab("ln(RR)")

```

# Look at LRR vs. amount added

```{r}
lta.data <- trt.data%>%
  rename(year = Year, Study_shorthand = Study) %>%
  mutate(year = as.character(year))

anpp.ppt <- anpp.4 %>%
  left_join(lta.data)

con <- anpp.ppt%>%
  filter(del == 0) %>%
  mutate(pairing = str_sub(trt, 1, 2)) %>%
  select(pairing, Study_shorthand, year, func, mean_ANPP, se_ANPP, Difference.from.LTA.MAP) %>%
  rename(ctrl_anpp = "mean_ANPP", 
         ctrl_se = "se_ANPP")

trts <- anpp.ppt  %>%
  filter(del==1) %>%
  mutate(pairing = str_sub(trt, 1, 2), 
         pairing = ifelse(pairing == "DR", "DR", "CO")) %>%
  select(pairing, Study_shorthand, year, func, mean_ANPP, se_ANPP,Size, Deluge.Timing) %>%
  rename(del_anpp = "mean_ANPP", 
         del_se = "se_ANPP")

paired <- left_join(con, trts)

paired.2 <- paired %>%
  mutate(lrr_anpp = log(del_anpp/ctrl_anpp), 
         anpp_diff = del_anpp - ctrl_anpp)

ggplot(paired.2, aes(Size, lrr_anpp, color = Difference.from.LTA.MAP))+
  geom_point(size = 2)+
  theme_bw()+
  ylab("ln(RR)")+
  xlab("Deluge Size (mm)")+
  facet_wrap(~func)+
  scale_color_gradientn(colors = c("#8B4513", "#D2691E", "#F5DEB3", "#87CEEB", "#1E90FF", "#00008B"))

ggplot(paired.2, aes(Size, anpp_diff, color = Difference.from.LTA.MAP))+
  geom_point(size = 2)+
  theme_bw()+
  ylab("ANPP Difference (Deluge-Control)")+
  xlab("Deluge Size (mm)")+
  facet_wrap(~func)+
  scale_color_gradientn(colors = c("#8B4513", "#D2691E", "#F5DEB3", "#87CEEB", "#1E90FF", "#00008B"))


```


# Look at ANPP vs. PPT
```{r}
str(anpp.4)
str(trt.data)
lta.data <- trt.data%>%
  rename(year = Year, Study_shorthand = Study) %>%
  mutate(year = as.character(year))

anpp.ppt <- anpp.4 %>%
  left_join(lta.data)

str(anpp.ppt)
ggplot(anpp.ppt, aes(Difference.from.LTA.MAP, mean_ANPP))+
  geom_point(aes(shape = as.factor(del), color = Study_shorthand), size = 2)+
  geom_smooth(se= F, method="lm", size = .9, color = "darkgrey", linetype="dashed")+
  theme_bw()+
  facet_wrap(~func)
  
```

# Which studies have negative/postiive/no response

```{r}
str(paired.2)

paired.2$pue <- paired.2$anpp_diff/paired.2$Size

func.anpp.diff <- paired.2 %>%
  # filter to not include fall
  #filter(Study_shorthand %in% c("Hajek2024", "Tooley_DRE")) %>%
  group_by(func) %>%
  summarize(mean_diff = mean(anpp_diff), se_diff = (sd(anpp_diff))/(sqrt(n()))) 
 

library(viridis)
ggplot() +
  geom_jitter(data = paired.2, aes(x = anpp_diff, y = func, color = Size, shape = func), 
           alpha = 0.8, size = 2.4) +
    geom_point(data = func.anpp.diff, aes(x = mean_diff, y = func, shape = func), size = 4) +
  geom_errorbar(data = func.anpp.diff, aes(x = mean_diff, y = func,
                                     xmin = mean_diff - se_diff, 
                                     xmax = mean_diff + se_diff),
                width = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_viridis_c(option = "mako", direction = -1)+
  theme_bw(base_size = 14) +
  ylab("Functional Group") +
  xlab("ANPP Difference (g/m2)")

ggplot(paired.2, aes(Size, pue, color = Difference.from.LTA.MAP))+
  geom_point() +
  geom_smooth(method="lm")+
  theme_bw()+
  facet_wrap(~func)

ggplot(paired.2, aes(pue, Difference.from.LTA.MAP, color = Size))+
  geom_point() +
  geom_smooth(method="lm")+
  theme_bw()+
  facet_wrap(~func)


response.summary <- paired.2 %>%
  mutate(response_type = case_when(
    del_anpp - del_se > ctrl_anpp + ctrl_se ~ "POSITIVE", 
    del_anpp + del_se < ctrl_anpp - ctrl_se ~ "NEGATIVE",
    TRUE ~ "NONE"
  )) %>%
  group_by(func, response_type) %>%
  summarize(n = n())


response.summary2 <- paired.2 %>%
  mutate(response_type = case_when(
    del_anpp  > ctrl_anpp  ~ "POSITIVE", 
    del_anpp  < ctrl_anpp  ~ "NEGATIVE",
    TRUE ~ "NONE"
  )) %>%
  group_by(func, response_type) %>%
  summarize(n = n())

```

# quick look at bnpp
```{r}
bnpp <- anpp %>%
  filter(!is.na(BNPP)) %>%
  group_by(Study_shorthand, trt) %>%
  summarize(mean_bnpp=mean(BNPP), se_bnpp = (sd(BNPP))/(sqrt(n())),
            mean_anpp=mean(ANPP), se_anpp = (sd(ANPP))/(sqrt(n()))) %>%
  mutate(del = as.numeric(ifelse(trt %in% c("CON", "DR"), "0", "1")))%>%
  ungroup()

con.b <- bnpp %>%
  filter(del == 0) %>%
  select(-c(trt, del)) %>%
  rename(ctrl_bnpp = "mean_bnpp", 
         ctrl_bnpp_se = "se_bnpp",
         ctrl_anpp2 = "mean_anpp", 
         ctrl_anpp_se2 = "se_anpp")

trts.b <- bnpp %>%
  filter(del == 1) %>%
  rename(del_bnpp = "mean_bnpp", 
         del_bnpp_se = "se_bnpp",
         del_anpp2 = "mean_anpp", 
         del_anpp_se2 = "se_anpp") %>%
  left_join(con.b) %>%
  mutate(bnpp_diff = del_bnpp-ctrl_bnpp,
         anpp_diff = del_anpp2 - ctrl_anpp2, 
         func = "ANPP")

ggplot(trts.b, aes(anpp_diff, bnpp_diff, color = Study_shorthand))+
  geom_point(size=3) +
  theme_bw()

```


# MAKE AN OVERALL figure that has the response summary (increase in NPP) for ANPP, BNPP, FOrb, and grass

```{r}
bnpp_prep <- trts.b %>%
  left_join(paired) %>%
  dplyr::select(c(Study_shorthand, trt, del_bnpp, del_bnpp_se, ctrl_bnpp, ctrl_bnpp_se, 
           pairing, year, Difference.from.LTA.MAP)) %>%
  rename(del_anpp = del_bnpp, 
         del_se = del_bnpp_se, 
         ctrl_anpp = ctrl_bnpp, 
         ctrl_se = ctrl_bnpp_se) %>%
  mutate(func = "BNPP")

glimpse(trt.data)
trt.data.2 <- trt.data %>%
  select(c(Study, Size, Year, trt))%>%
  rename(Study_shorthand = Study, 
         year = Year)%>%
  mutate(year = as.character(year))
  

combined <- paired %>%
  rbind(bnpp_prep) %>%
  mutate(npp_diff = del_anpp-ctrl_anpp) %>%
  left_join(trt.data.2)%>%
  mutate(pue = npp_diff/Size, wyr_ppt = Size+ Difference.from.LTA.MAP)

combo_summary <- combined %>%
  group_by(func)%>%
  summarize(mean_npp = mean(npp_diff), se_npp = (sd(npp_diff))/(sqrt(n())))

combo_summary$func <- factor(combo_summary$func, levels = c("BNPP", "FORB", "GRASS", "ANPP"))

ggplot() +
  geom_jitter(data = combined, aes(x = npp_diff, y = func, fill = Size, shape = func), 
           alpha = 0.65, size = 2.3) +
    geom_point(data = combo_summary, aes(x = mean_npp, y = func, shape = func), size = 4.5, fill = "black") +
  geom_errorbar(data = combo_summary, aes(x = mean_npp, y = func,
                                     xmin = mean_npp - se_npp, 
                                     xmax = mean_npp + se_npp),
                width = 0.2) +
  scale_shape_manual(values = c(23, 24, 22, 21))+
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_viridis_c(option = "mako", direction = -1)+
  theme_bw(base_size = 14) +
  ylab("Functional Group") +
  xlab("ANPP Difference (g/m2)")

ggplot(data = combined, aes(del_ppt, pue)) +
  geom_point(aes(color = Study_shorthand ))+
  theme_bw()+
  geom_smooth(method="lm")+
  facet_wrap(~func)


```

# Comparison of slopes
## Compare with controls of treatments and deluge treatments - maybe just the 60mm ones?
## Bring in the overall CPER data

```{r}
anpp.ppt.2 <- anpp.ppt %>%
  mutate(MAP = Difference.from.LTA.MAP + 333) %>%
  filter(func == "ANPP") %>%
  # filter(!(Study_shorthand == "Tooley_DRE" & trt %in% c("DR", "DR_DEL"))) #%>%
  filter(Size== 60)

glimpse(anpp.ppt.2)

ggplot(anpp.ppt.2, aes(MAP, mean_ANPP, color = as.factor(del))) +
  geom_point()+
  geom_smooth(method = "lm")+
  theme_bw()+
  facet_wrap(~func)

ggplot(anpp.ppt.2, aes(MAP, mean_ANPP, color = Study_shorthand)) +
  geom_point()+
  theme_bw()


# read in CPER data
cper <- read.csv("/Users/olhajek/Desktop/nceas/lterwg-resilience/data/harmonized_data/anpp_wyr_merged.csv") %>%
  filter(site=="CPER")

ggplot()+
  geom_point(data = anpp.ppt.2, aes(MAP, mean_ANPP, color = as.factor(del))) +
  geom_smooth(data = anpp.ppt.2, aes(MAP, mean_ANPP, color = as.factor(del)), method = "lm")+
  geom_point(data=cper, aes(wyr_ppt, anpp_g_m2))+
  geom_smooth(data=cper, aes(wyr_ppt, anpp_g_m2), method = "lm")+
  theme_bw() 

```


# Create a df for anpp

```{r}
glimpse(anpp.4)
# create an all treatments with anpp.4
trt.data.3 <- trt.data %>%
  select(c(Study, Size, Year, trt, Difference.from.LTA.MAP))%>%
  rename(Study_shorthand = Study, 
         year = Year)%>%
  mutate(year = as.character(year))

anpp.5 <- anpp.4 %>%
  left_join(trt.data.3) %>%
  mutate(wyr_ppt = Difference.from.LTA.MAP + 333)

# save the combined and the anpp

write.csv(x = anpp.5, row.names = F, na = '',
          file = file.path("deluge", "data", "R_exports","ANPP_treatments_aggregated.csv"))

# Upload to Drive
drive_upload(
  media = file.path("deluge", "data", "R_exports", "ANPP_treatments_aggregated.csv"),
  path = file.path("deluge", "data","R_exports"),
  name = "ANPP_treatments_aggregated.csv",
  overwrite = T)

write.csv(x = combined, row.names = F, na = '',
          file = file.path("deluge", "data", "R_exports","ANPP_paired_aggregated.csv"))

# Upload to Drive
drive_upload(
  media = file.path("deluge", "data", "R_exports", "ANPP_paired_aggregated.csv"),
  path = file.path("deluge", "data","R_exports"),
  name = "ANPP_paired_aggregated.csv",
  overwrite = T)


```



