---
title: "ESA_continuous_analysis"
author: "Kathy Condon"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(tidyverse)
library(dplyr)
library(car)
library(emmeans)
library(multcomp)
library(lmerTest)
library(lme4)
library(RColorBrewer)
library(rstatix)
library(tidyr)
library(ggpubr)
library(lubridate)
library(ggtext)
library(ggh4x)
library(ggpattern)
library(data.table)
library(MuMIn)
library(scales)
library(performance)
library(googledrive)

```

## Plan for Continuous data in the talk:
1. Filter out Mary totally, filter out Greg drought legacies
2. Use longest response period from published (probably Alison big deluge) and find maximums to check, and AUC to analyze more (potentially split also into immediate and full response periods)
3. AUC ~ timing*size for everyone to look for those being important [[update for 20 July after looking at this -- likly not enough replication in each to look at this formally w regression]] -- do just with growing season vs. fall timing and normal vs. extra large sizes (80mm??) 
4. Probably make some bar plots of AUCs to compare magnitude (for pre-/response (or response split)/post-) -- maybe keep fall out of post if that doesn't make sense?
5. Add relationship of GCC to ANPP -- ANPP increase w/ deluge ~ GCC's AUC (OR change to timing like relative 20%) (OR ANPP's difference explained by BOTH GCC magnitude over consistent periods AND GCC total increased length within each study)

Probably greenness and soil moisture (only Resp. if it's very interesting)

### Data setup
```{r read data}
# set local file directories if needed
dir.create(file.path("deluge"), showWarnings = F)
dir.create(file.path("deluge","data"), showWarnings = F)
dir.create(file.path("deluge","data","R_exports"))
dir.create(file.path("deluge","figures"))

# read in continuous files
drive.files<-googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3?usp=drive_link"))

cont.files<-c("GCC_filtered.csv","SR_filtered.csv","VWC_20cm_filtered.csv","deluge_characteristics_mergeable.csv","deluge_trt_keys.csv")

## read in each file
for (i in cont.files){
  
file.name<-drive.files %>% 
  dplyr::filter(name == i) 

# if-else, to throw error if file name is incorrect or error would otherwise stop loop
if(nrow(file.name) == 1){
  googledrive::drive_download(file = file.name$id, 
                              overwrite = T,
                              path = file.path("deluge", 'data', i))
  
} else{
  message("File error - check name or if duplicated: ", i)
  }
}
```

```{r check data}
# check all continuous files
## read in
GCC<-read.csv("deluge/data/GCC_filtered.csv")
str(GCC) # 6103 -- OK

VWC<-read.csv("deluge/data/VWC_20cm_filtered.csv")
str(VWC) # 6323 -- OK

SR<-read.csv("deluge/data/SR_filtered.csv")
str(SR) # 3753 -- OK

## check each
# GCC
tapply(GCC$year, GCC$Study_shorthand, unique)

GCC.dat<-GCC%>%
  mutate(var=as.numeric(GCC))%>%
  filter(!Study_shorthand=="Linabury_ChANGE")%>% # drop Mary 
  filter(!(Study_shorthand=="Tooley_DRE" & trt %in% c("DR","DR_DEL"))) # only keep Greg's control (non-drought legacy) plots

GCC.na<-GCC.dat%>%
  filter(is.na(var))
# view(GCC.na)

tapply(GCC.dat$year, GCC.dat$Study_shorthand, unique)
unique(GCC.dat$Study_shorthand)

# quiz viz of all #
ggplot(GCC.dat, aes(date, var))+
  geom_point()+
  theme_bw()+
  facet_wrap(~Study_shorthand, scales="free") # OK- nothing looks too unreasonable

# VWC
tapply(VWC$year, VWC$Study_shorthand, unique)

VWC.dat<-VWC%>%
  mutate(var=as.numeric(VWC))%>%
  filter(!Study_shorthand=="Linabury_ChANGE")%>% # drop Mary 
  filter(!(Study_shorthand=="Tooley_DRE" & trt %in% c("DR","DR_DEL"))) # only keep Greg's control (non-drought legacy) plots

VWC.na<-VWC.dat%>%
  filter(is.na(var))
# view(VWC.na)

# quiz viz of all #
ggplot(VWC.dat, aes(date, var))+
  geom_point()+
  theme_bw()+
  facet_wrap(~Study_shorthand, scales="free") # OK- nothing looks too unreasonable

# check 
tapply(VWC$year, VWC$Study_shorthand, unique)

# SR
tapply(SR$year, SR$Study_shorthand, unique)

SR.dat<-SR%>%
  mutate(var=as.numeric(efflux))%>%
  filter(!(Study_shorthand=="Tooley_DRE" & trt %in% c("DR","DR_DEL")))

SR.na<-SR.dat%>%
  filter(is.na(var))
# view(SR.na)

# quiz viz of all #
ggplot(SR.dat, aes(date, var))+
  geom_point()+
  theme_bw()+
  facet_wrap(~Study_shorthand, scales="free") # OK- nothing looks too unreasonable

# check
tapply(SR$year, SR$Study_shorthand, unique)
```

```{r list checked/fixed files}
# List together
con.dat<-list(VWC.dat,GCC.dat,SR.dat)
con.labels<-list("20 cm VWC (%)","GCC","Efflux")

## also read in deluge characteristics data to merge with
del.info<-read.csv("deluge/data/deluge_characteristics_mergeable.csv")
# view(del.info)
str(del.info)

del.info<-del.info%>%
  mutate(Study_trt=paste0(Study_shorthand,"_",trt))

## and trt key by study/year
del.trts<-read.csv("deluge/data/deluge_trt_keys.csv")%>%
  rename(control_trt=con)
str(del.trts)
del.trts
```

### Assign response periods
```{r response period assignment}
#### Post -- 28 days response period in both studies -- using approximately when largest deluge VWC returned to ambient levels

##### assign response periods ######
##### from del.trts -- which already duplicates controls where needed for multiple deluge treatments
### response period info 
resp.assign<-del.trts%>%
  # format dates for deluge start and end and create response period end date (+28 days after start)
  mutate(del_start=as.Date(del_start,format="%m/%d/%y"),
         del_end=as.Date(del_end,format="%m/%d/%y"),
         resp_period_end=del_start+28)

### add in deluge info from del.info
trt.pairs<-resp.assign%>% # keep all from resp.assign
  left_join(del.info%>%select(Study_shorthand,year,trt,del_size,timing), # add these columns from del.info file too
            by=c("Study_shorthand","year","trt"))

# ### also remove extra DRE trts
# trt.pairs<-trt.pairs %>%
#   filter(!(Study_shorthand == "Tooley_DRE" & trt %in% c("DR_DEL"))) # remove Tooley DR legacy plots, only keep the deluged plots w/ ambient history

## stack treated and control rows together with their date windows (and duplicate controls where needed for multiple deluge trts)
trt.lookup.long<-trt.pairs%>%
  select(Study_shorthand,year,trt,control_trt,del_size,timing,del_start,del_end,resp_period_end)%>%
  pivot_longer(cols=c(trt,control_trt),names_to="con_vs_trt",values_to="trt")

head(trt.lookup.long)

##### join response periods w/ full continuous data ######
## assign each continuous var's data to mark response periods by study and trt
cont.list<-list(GCC=GCC.dat,SR=SR.dat,VWC=VWC.dat)

## join response dates into continuous data measurements
for (i in seq_along(cont.list)) {
  cont.list[[i]]<-cont.list[[i]]%>%
    mutate(date=as.Date(date,format="%Y-%m-%d"))%>%
    left_join(trt.lookup.long,by=c("Study_shorthand","year","trt"),relationship="many-to-many")%>%
    mutate(
      resp_period=case_when(
        date <= del_start ~ "pre_del",
        date > del_start & date <= resp_period_end ~ "del_resp",
        date > resp_period_end ~ "post_del",
        TRUE ~ NA_character_)) # NA's for when there are not del_start/del_end/resp_period_end dates -- primarily for years w/ fall deluges that continue into next year (& next year will not have any deluge dates)
}

GCC.dat.new<-cont.list$GCC
SR.dat.new<-cont.list$SR
VWC.dat.new<-cont.list$VWC
```

### check data
```{r combine paired trt-controls data}
#### check data # of rows, trt combos, etc. ####
# "new" dats should be slightly more n b/c some controls have duplicated to match all trts
nrow(GCC.dat)
nrow(GCC.dat.new)
nrow(GCC.dat.new)/nrow(GCC.dat)

nrow(SR.dat)
nrow(SR.dat.new)
nrow(SR.dat.new)/nrow(SR.dat)

nrow(VWC.dat)
nrow(VWC.dat.new)
nrow(VWC.dat.new)/nrow(VWC.dat)

# check how many pts per del period
# dat<-GCC.dat.new
# dat<-SR.dat.new
dat<-VWC.dat.new

point.counts<-dat%>%
  dplyr::count(Study_shorthand, year, trt, resp_period)%>% # how many data points are in eacy period
  arrange(resp_period,Study_shorthand)
point.counts

# check any studies missing boundaries
dat%>%
  group_by(Study_shorthand,year,trt)%>%
  summarise(
    del_start=first(del_start),
    resp_period_end=first(resp_period_end),
    first_date=min(date),
    last_date=max(date),
    n=n())%>%
  mutate(missing_boundary=first_date>del_start | last_date<resp_period_end) %>%
  filter(missing_boundary=="TRUE")

#### merge all continuous data together ####
str(VWC.dat.new)
str(GCC.dat.new)
str(SR.dat.new)

VWC.tomerge<-VWC.dat.new%>%
  select(-c(VWC))

GCC.tomerge<-GCC.dat.new%>%
  select(-c(GCC))

SR.tomerge<-SR.dat.new%>%
  select(-c(efflux))%>%
  mutate(plot=as.character(plot))

all.continuous<-bind_rows(VWC.tomerge,GCC.tomerge,SR.tomerge)%>%
  mutate(date=as.Date(date,format="%Y-%m-%d"),
         date=if_else(as.numeric(format(date,"%Y"))<100,
                      date+years(2000),
                      date))
str(all.continuous)

#### averages and plot checks ####
cont.avs<-all.continuous%>%
  group_by(Study_shorthand,Variable,year,date,trt,con_vs_trt,del_size,timing,del_start,del_end,resp_period_end,resp_period)%>%
  dplyr::summarise(
    n.plots.per.date=n(),
    trt.day.av=mean(var,na.rm=TRUE),
    trt.day.SE=(sd(var,na.rm=TRUE))/sqrt(n.plots.per.date))%>%
  filter(!(Study_shorthand=="Tooley_DRE" & year == "2026"))
# view(cont.avs)

## try plotting to check data
### take averages
dat<-cont.avs%>%
  # subset(Variable=="VWC")%>%
  subset(Variable=="GCC")%>%
  # subset(Variable=="SR")%>%
  group_by(Study_shorthand,year)%>%
  mutate(Study_trt=paste0(Study_shorthand,"_",trt),
         trt_type=factor(con_vs_trt,levels=c("control_trt","trt"),labels=c("CON","DEL")))

### specify response period shading
resp_rects<-dat%>%
  filter(!is.na(del_end),!is.na(resp_period_end))%>%
  distinct(Study_shorthand,year,del_start,resp_period_end)
  
### make the plot
plot<-ggplot(dat,aes(x=date,y=trt.day.av,group=trt,linetype=trt_type,color=Study_trt)) +
  geom_rect(data = resp_rects,aes(xmin = del_start,xmax = resp_period_end,ymin = -Inf,ymax = Inf),
            inherit.aes = FALSE,
            fill = "grey80",
            alpha = 0.3)+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=trt.day.av-trt.day.SE, ymax=trt.day.av+trt.day.SE), width=0.2)+
  scale_x_date(date_labels="%b %d") +
  labs(
    x="Date",
    y="VWC",
    linetype="Treatment")+
  theme_bw()+
  theme(legend.position="right")+
  force_panelsizes(rows = unit(2.25, "in"),
               cols = unit(4.25, "in"))+
  facet_wrap(~Study_shorthand + year,scales="free")

plot
```

### Calc. AUC's for all
```{r PREP AUC's - magnitude}
## load package for trapazoidal AUC calculation 
library(pracma)

#### Approximate for AUC boudaries ####
### get daily averages
daily.avs<-all.continuous%>%
  group_by(Variable,Study_shorthand,year,trt,date,del_size,timing,del_start,del_end,resp_period_end,resp_period)%>%
  dplyr::summarise(
    n.plots.per.date=sum(!is.na(var)),
    trt.av=mean(var,na.rm=TRUE),
    trt.sd=sd(var,na.rm=TRUE),
    trt.SE=trt.sd/sqrt(n.plots.per.date))

### check studies missing observations before/after boundaries
missing.bounds<-daily.avs%>%
  group_by(Variable,Study_shorthand,year,trt)%>%
  summarise(
    del_start=first(del_start), # first() just grabs the first row's date since all rows repeat the date
    resp_period_end=first(resp_period_end),
    first_date=min(date),
    last_date=max(date),
    n=n())%>%
  mutate(missing_boundary=
           # observations start after deluge started
           first_date>del_start | # OR
           # observations ended before deluge response period ended
           last_date<resp_period_end) %>%
  filter(missing_boundary=="TRUE")
view(missing.bounds)
  
#### interpolate missing AUC boundaries ####
#### for calculating response period AUC's
auc.dat<-daily.avs%>%
  # drop Tooley GCC 2023-2024 b/c missing boundaries
  filter(!(Study_shorthand=="Tooley_DRE" & Variable == "GCC" & year %in% c(2023,2024)))%>%
  # drop Tooley SR 2024 b/c missing boundaries
  filter(!(Study_shorthand=="Tooley_DRE" & Variable == "SR" & year == "2024"))%>%
  filter(!is.na(del_start),!is.na(resp_period_end))


#### check that grouping combo only have gives one del_start/resp_period_end
check<-auc.dat%>%
  dplyr::count(Variable,Study_shorthand,year,trt,del_start,resp_period_end)%>%
  filter(n > 1)
check # want to be nothing

#### interpolating and adding points on boundaries
## interpolate
auc.dat.bounds<-auc.dat%>%
  group_by(Variable,Study_shorthand,year,trt,del_size,timing,del_start,del_end,resp_period_end)%>%
  # arrange date to put in order (matters for as.numeric(date))
  arrange(date)%>%
  dplyr::summarise(
    start_point=approx(
      x=as.numeric(date),
      y=trt.av,
      xout=as.numeric(first(del_start))
    )$y,
    end_point=approx(
      x=as.numeric(date),
      y=trt.av,
      xout=as.numeric(first(resp_period_end))
    )$y)

# view(auc.dat.bounds)

## make longer (single pt per date)
auc.dat.bounds.long<-auc.dat.bounds%>%
  pivot_longer(cols=c(start_point,end_point),
               names_to="side",
               values_to="trt.av")%>%
  # assign single date to interpolated value for boundary date
  mutate(date=case_when(
    side=="start_point"~del_start,
    side=="end_point"~resp_period_end))%>%
  select(!side)

# view(auc.dat.bounds.long)

## "anti-join" to ONLY keep the interpolated point (auc.dat.bounds.long) if the auc daily avs (auc.dat) do not have a point there already -- so if someone did measure on an exact boundary day, we don't interpolate
auc.new.bounds<-auc.dat.bounds.long%>%
  anti_join(auc.dat,
            by=c("Variable","Study_shorthand","year","trt","del_size","timing","del_start","del_end","resp_period_end","date"))

# view(auc.new.bounds)
nrow(auc.dat.bounds.long) # 184 total interpolated points
nrow(auc.new.bounds) # 116 left when we drop where someone did measure on those dates

## add to the whole data
nrow(auc.dat) # 2030

auc.dat.filled<-bind_rows(
  auc.dat%>%mutate(interpolated=FALSE),
  auc.new.bounds%>%mutate(interpolated=TRUE))%>% # combine
  group_by(Variable,Study_shorthand,year,trt)%>% # sort nicely again
  arrange(date,.by_group=TRUE)%>% # and put dates in order
  ungroup()

nrow(auc.dat.filled) # 2146 == 2030 + 116
```

```{r CHECK AND CALC AUC's - magnitude}
#### take averages
dat<-auc.dat.filled%>%
  # subset(Variable=="VWC")%>%
  # subset(Variable=="GCC")%>%
  subset(Variable=="SR")%>%
  group_by(Study_shorthand,year)%>%
  mutate(Study_trt=paste0(Study_shorthand,"_",trt),
         trt_type=case_when(
           trt %in% c("CON","DR") ~ "CON",
           TRUE ~ "DEL"))%>%
  filter(Study_shorthand %in% c("Hajek2024","Post2020"))

### specify response period shading
resp_rects<-dat%>%
  filter(!is.na(del_end),!is.na(resp_period_end))%>%
  distinct(Study_shorthand,year,del_start,resp_period_end)
  
### make the plot
filled.plot<-ggplot(dat,aes(x=date,y=trt.av,group=trt,linetype=trt_type,color=Study_trt)) +
  geom_rect(data = resp_rects,aes(xmin = del_start,xmax = resp_period_end,ymin = -Inf,ymax = Inf),
            inherit.aes = FALSE,
            fill = "grey80",
            alpha = 0.3)+
  geom_line()+
  geom_point(aes(shape=interpolated), size=3)+
  geom_errorbar(aes(ymin=trt.av-trt.SE, ymax=trt.av+trt.SE), width=0.2)+
  scale_x_date(date_labels="%b %d") +
  labs(
    x="Date",
    y="VWC",
    linetype="Treatment")+
  theme_bw()+
  theme(legend.position="right")+
  force_panelsizes(rows = unit(2.25, "in"),
               cols = unit(4.25, "in"))+
  facet_wrap(~Study_shorthand + year,scales="free")

filled.plot

#### calculate auc's ####
auc.calcs<-auc.dat.filled%>%
  # filter to only dates within the response period based on each row's date and del/resp_period dates
  filter(date>=del_start & date<=resp_period_end)%>%
  group_by(Variable,Study_shorthand,year,trt,del_size,timing,del_start,del_end,resp_period_end)%>%
  # make sure dates are in order
  arrange(date,.by_group=TRUE)%>%
  # calculate!
  dplyr::summarise(
    AUC=trapz(as.numeric(date),trt.av)
  )

auc.calcs.paired <- auc.calcs %>%
  left_join(trt.pairs %>% select(Study_shorthand, year, trt, control_trt, del_start, resp_period_end),
            by = c("Study_shorthand", "year", "trt", "del_start", "resp_period_end"))

# separate trts and controls
auc.trts <- auc.calcs.paired %>%
  filter(!is.na(control_trt))  # only deluge treatments have a control_trt

auc.cons <- auc.calcs.paired %>%
  filter(is.na(control_trt)) %>%
  ungroup() %>%
  select(Variable, Study_shorthand, year, trt, del_start, resp_period_end, AUC) %>%
  rename(AUC_con = AUC, control_trt = trt) %>%
  distinct()

# join by matching control_trt to the control's trt name
auc.diff <- auc.trts %>%
  ungroup() %>%
  left_join(auc.cons,
            by = c("Variable", "Study_shorthand", "year", "control_trt", "del_start", "resp_period_end")) %>%
  mutate(
    AUC_diff     = AUC - AUC_con,
    AUC_day      = AUC / 28,
    AUC_con_day  = AUC_con / 28,
    AUC_diff_day = AUC_diff / 28
  )

view(auc.diff)

## check 
ggplot(auc.diff, aes(x = Study_shorthand, y = AUC_diff_day, color = factor(del_size), shape = timing)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Variable, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(auc.diff, aes(x = del_size, y = AUC_diff_day, color = Study_shorthand, shape = timing)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Variable, scales = "free_y") +
  theme_bw()

ggplot(auc.diff, aes(x = del_size, y = AUC_diff_day, color = Study_shorthand)) +
  geom_point(size = 3, aes(shape = timing)) +
  geom_smooth(method = "lm", se = TRUE, aes(group = 1), color = "black", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~Variable, scales = "free_y") +
  theme_bw()
```



```{r rel 20% - resp lengths}


## 20% stuff again



```


```{r period averages}
## average (pre-deluge, post-deluge through the end of year, response period)


```


















```{r magnitude bar graphs}
# ggplot(GCC.auc, aes(x = del_size, y = AUC_diff, color = Study_shorthand, shape = timing)) +
#   geom_point(size = 3) +
#   geom_smooth(method = "lm", se = FALSE, aes(group = 1), color = "black", linetype = "dashed") +
#   labs(x = "Deluge Size (mm)", y = "AUC Difference (Treated - Control)",
#        title = "GCC Response Period AUC vs. Deluge Size",
#        color = "Study", shape = "Timing") +
#   theme_classic()
# 
# ggplot(GCC.auc, aes(x = timing, y = AUC_diff, color = Study_shorthand)) +
#   geom_point(size = 3, position = position_jitter(width = 0.1)) +
#   labs(x = "Timing", y = "AUC Difference (Treated - Control)",
#        title = "GCC Response Period AUC vs. Deluge Timing",
#        color = "Study") +
#   theme_classic()

### summarize all together
# get all unique treatment combinations across all three variables
all.trts <- bind_rows(
  GCC.auc %>% select(Study_shorthand, year, trt, del_size, timing),
  VWC.auc %>% select(Study_shorthand, year, trt, del_size, timing),
  SR.auc  %>% select(Study_shorthand, year, trt, del_size, timing)
) %>% distinct()

# join all three variables to the full list
auc.summary <- all.trts %>%
  left_join(GCC.auc %>% select(Study_shorthand, year, trt, del_size, timing, AUC_diff) %>% rename(AUC_diff_GCC = AUC_diff),
            by = c("Study_shorthand", "year", "trt", "del_size", "timing")) %>%
  left_join(VWC.auc %>% select(Study_shorthand, year, trt, del_size, timing, AUC_diff) %>% rename(AUC_diff_VWC = AUC_diff),
            by = c("Study_shorthand", "year", "trt", "del_size", "timing")) %>%
  left_join(SR.auc  %>% select(Study_shorthand, year, trt, del_size, timing, AUC_diff) %>% rename(AUC_diff_SR  = AUC_diff),
            by = c("Study_shorthand", "year", "trt", "del_size", "timing"))

# look at highest differences by each var
auc.summary%>%
  arrange(desc(AUC_diff_VWC))

auc.summary%>%
  arrange(desc(AUC_diff_GCC))

auc.summary%>%
  arrange(desc(AUC_diff_SR))

auc.summary.bydays<-auc.summary %>%
  mutate(across(starts_with("AUC_diff"), ~ . / 28)) %>%
  kable(digits = 3,
        col.names = c("Study", "Year", "Treatment", "Deluge Size (mm)", "Timing",
                      "GCC mean diff", "VWC mean diff (%)", "SR mean diff"),
        caption = "Mean daily response above control during 28-day response period")

# look at by study ranks all together
auc.summary %>%
  mutate(across(starts_with("AUC_diff"), ~ . / 28)) %>%
  mutate(
    rank_GCC = rank(-AUC_diff_GCC),
    rank_VWC = rank(-AUC_diff_VWC),
    rank_SR  = rank(-AUC_diff_SR)
  ) %>%
  select(Study_shorthand, year, trt, del_size, timing, 
         rank_GCC,rank_VWC,rank_SR)

```


### AUC ~ timing and size regressions 
```{r magnitude regressions}


```

### ANPP explained by GCC magnitude/timing?
```{r get ANPP data setup}


```

```{r ANPP and GCC relationships}


```

### VPD maybe? vs. VWC?? vs. GCC?? vs. ANPP??
How much does VWC increase depends on VPD

