---
title: "ESA_continuous_analysis"
author: "Kathy Condon"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(tidyverse)
library(dplyr)
library(car)
library(emmeans)
library(multcomp)
library(lmerTest)
library(lme4)
library(RColorBrewer)
library(rstatix)
library(tidyr)
library(ggpubr)
library(lubridate)
library(ggtext)
library(ggh4x)
library(ggpattern)
library(data.table)
library(MuMIn)
library(scales)
library(performance)
library(googledrive)

```

## Plan for Continuous data in the talk:
1. Filter out Mary totally, filter out Greg drought legacies
2. Use longest response period from published (probably Alison big deluge) and find maximums to check, and AUC to analyze more (potentially split also into immediate and full response periods)
3. AUC ~ timing*size for everyone to look for those being important
4. Probably make some bar plots of AUCs to compare magnitude (for pre-/response (or response split)/post-) -- maybe keep fall out of post if that doesn't make sense?
5. Add relationship of GCC to ANPP -- ANPP increase w/ deluge ~ GCC's AUC (OR change to timing like relative 20%) (OR ANPP's difference explained by BOTH GCC magnitude over consistent periods AND GCC total increased length within each study)

Probably greenness and soil moisture (only Resp. if it's very interesting)

### Data setup
```{r data}




```

### Assign response periods
```{r response period assignment}



```

### Calc. AUC's for all
```{r AUC's}



```

### check AUC's compared over pre-, response periods, and post-del
```{r magnitude bar graphs}


```

### AUC ~ timing and size regressions 
```{r magnitude regressions}


```

### ANPP explained by GCC magnitude/timing?
```{r get ANPP data setup}


```

```{r ANPP and GCC relationships}


```

