---
title: "ESA_continuous_analysis"
author: "Kathy Condon"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(tidyverse)
library(dplyr)
library(car)
library(emmeans)
library(multcomp)
library(lmerTest)
library(lme4)
library(RColorBrewer)
library(rstatix)
library(tidyr)
library(ggpubr)
library(lubridate)
library(ggtext)
library(ggh4x)
library(ggpattern)
library(data.table)
library(MuMIn)
library(scales)
library(performance)
library(googledrive)

```

## Plan for Continuous data in the talk:
1. Filter out Mary totally, filter out Greg drought legacies
2. Use longest response period from published (probably Alison big deluge) and find maximums to check, and AUC to analyze more (potentially split also into immediate and full response periods)
3. AUC ~ timing*size for everyone to look for those being important [[update for 20 July after looking at this -- likly not enough replication in each to look at this formally w regression]] -- do just with growing season vs. fall timing and normal vs. extra large sizes (80mm??) 
4. Probably make some bar plots of AUCs to compare magnitude (for pre-/response (or response split)/post-) -- maybe keep fall out of post if that doesn't make sense?
5. Add relationship of GCC to ANPP -- ANPP increase w/ deluge ~ GCC's AUC (OR change to timing like relative 20%) (OR ANPP's difference explained by BOTH GCC magnitude over consistent periods AND GCC total increased length within each study)

Probably greenness and soil moisture (only Resp. if it's very interesting)

### Data setup
```{r read data}
# set local file directories if needed
dir.create(file.path("deluge"), showWarnings = F)
dir.create(file.path("deluge","data"), showWarnings = F)
dir.create(file.path("deluge","data","R_exports"))
dir.create(file.path("deluge","figures"))

# read in continuous files
drive.files<-googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3?usp=drive_link"))

cont.files<-c("GCC_filtered.csv","SR_filtered.csv","VWC_20cm_filtered.csv","deluge_characteristics_mergeable.csv","deluge_trt_keys.csv")

## read in each file
for (i in cont.files){
  
file.name<-drive.files %>% 
  dplyr::filter(name == i) 

# if-else, to throw error if file name is incorrect or error would otherwise stop loop
if(nrow(file.name) == 1){
  googledrive::drive_download(file = file.name$id, 
                              overwrite = T,
                              path = file.path("deluge", 'data', i))
  
} else{
  message("File error - check name or if duplicated: ", i)
  }
}
```

```{r check data}
# check all continuous files
## read in
GCC<-read.csv("deluge/data/GCC_filtered.csv")
str(GCC) # 6103 -- OK

VWC<-read.csv("deluge/data/VWC_20cm_filtered.csv")
str(VWC) # 6323 -- OK

SR<-read.csv("deluge/data/SR_filtered.csv")
str(SR) # 3753 -- OK

## check each
# GCC
tapply(GCC$year, GCC$Study_shorthand, unique)

GCC.dat<-GCC%>%
  mutate(var=as.numeric(GCC))%>%
  filter(!Study_shorthand=="Linabury_ChANGE")%>% # drop Mary 
  filter(!(Study_shorthand=="Tooley_DRE" & trt %in% c("DR","DR_DEL"))) # only keep Greg's control (non-drought legacy) plots

GCC.na<-GCC.dat%>%
  filter(is.na(var))
# view(GCC.na)

tapply(GCC.dat$year, GCC.dat$Study_shorthand, unique)
unique(GCC.dat$Study_shorthand)

# quiz viz of all #
ggplot(GCC.dat, aes(date, var))+
  geom_point()+
  theme_bw()+
  facet_wrap(~Study_shorthand, scales="free") # OK- nothing looks too unreasonable

# VWC
tapply(VWC$year, VWC$Study_shorthand, unique)

VWC.dat<-VWC%>%
  mutate(var=as.numeric(VWC))%>%
  filter(!Study_shorthand=="Linabury_ChANGE")%>% # drop Mary 
  filter(!(Study_shorthand=="Tooley_DRE" & trt %in% c("DR","DR_DEL"))) # only keep Greg's control (non-drought legacy) plots

VWC.na<-VWC.dat%>%
  filter(is.na(var))
# view(VWC.na)

# quiz viz of all #
ggplot(VWC.dat, aes(date, var))+
  geom_point()+
  theme_bw()+
  facet_wrap(~Study_shorthand, scales="free") # OK- nothing looks too unreasonable

# check 
tapply(VWC$year, VWC$Study_shorthand, unique)

# SR
tapply(SR$year, SR$Study_shorthand, unique)

SR.dat<-SR%>%
  mutate(var=as.numeric(efflux))%>%
  filter(!(Study_shorthand=="Tooley_DRE" & trt %in% c("DR","DR_DEL")))

SR.na<-SR.dat%>%
  filter(is.na(var))
# view(SR.na)

# quiz viz of all #
ggplot(SR.dat, aes(date, var))+
  geom_point()+
  theme_bw()+
  facet_wrap(~Study_shorthand, scales="free") # OK- nothing looks too unreasonable

# check
tapply(SR$year, SR$Study_shorthand, unique)
```

```{r list checked/fixed files}
# List together
con.dat<-list(VWC.dat,GCC.dat,SR.dat)
con.labels<-list("20 cm VWC (%)","GCC","Efflux")

## also read in deluge characteristics data to merge with
del.info<-read.csv("deluge/data/deluge_characteristics_mergeable.csv")
# view(del.info)
str(del.info)

del.info<-del.info%>%
  mutate(Study_trt=paste0(Study_shorthand,"_",trt))

## and trt key by study/year
del.trts<-read.csv("deluge/data/deluge_trt_keys.csv")%>%
  rename(control_trt=con)
str(del.trts)
del.trts
```

### Assign response periods
```{r response period assignment}
#### Post -- 28 days response period in both studies -- using approximately when largest deluge VWC returned to ambient levels

##### assign response periods ######
##### from del.trts -- which already duplicates controls where needed for multiple deluge treatments
### response period info 
resp.assign<-del.trts%>%
  # format dates for deluge start and end and create response period end date (+28 days after start)
  mutate(del_start=as.Date(del_start,format="%m/%d/%y"),
         del_end=as.Date(del_end,format="%m/%d/%y"),
         resp_period_end=del_start+28)

### add in deluge info from del.info
trt.pairs<-resp.assign%>% # keep all from resp.assign
  left_join(del.info%>%select(Study_shorthand,year,trt,del_size,timing), # add these columns from del.info file too
            by=c("Study_shorthand","year","trt"))

# ### also remove extra DRE trts
# trt.pairs<-trt.pairs %>%
#   filter(!(Study_shorthand == "Tooley_DRE" & trt %in% c("DR_DEL"))) # remove Tooley DR legacy plots, only keep the deluged plots w/ ambient history

## stack treated and control rows together with their date windows (and duplicate controls where needed for multiple deluge trts)
trt.lookup.long<-trt.pairs%>%
  select(Study_shorthand,year,trt,control_trt,del_size,timing,del_start,del_end,resp_period_end)%>%
  pivot_longer(cols=c(trt,control_trt),names_to="con_vs_trt",values_to="trt")

head(trt.lookup.long)

##### join response periods w/ full continuous data ######
## assign each continuous var's data to mark response periods by study and trt
cont.list<-list(GCC=GCC.dat,SR=SR.dat,VWC=VWC.dat)

## join response dates into continuous data measurements
for (i in seq_along(cont.list)) {
  cont.list[[i]]<-cont.list[[i]]%>%
    mutate(date=as.Date(date,format="%Y-%m-%d"))%>%
    left_join(trt.lookup.long,by=c("Study_shorthand","year","trt"),relationship="many-to-many")%>%
    mutate(
      resp_period=case_when(
        date <= del_start ~ "pre_del",
        date > del_start & date <= resp_period_end ~ "del_resp",
        date > resp_period_end ~ "post_del",
        TRUE ~ NA_character_)) # NA's for when there are not del_start/del_end/resp_period_end dates -- primarily for years w/ fall deluges that continue into next year (& next year will not have any deluge dates)
}

GCC.dat.new<-cont.list$GCC
SR.dat.new<-cont.list$SR
VWC.dat.new<-cont.list$VWC
```

### check data
```{r combine paired trt-controls data}
#### check data # of rows, trt combos, etc. ####
# "new" dats should be slightly more n b/c some controls have duplicated to match all trts
nrow(GCC.dat)
nrow(GCC.dat.new)
nrow(GCC.dat.new)/nrow(GCC.dat)

nrow(SR.dat)
nrow(SR.dat.new)
nrow(SR.dat.new)/nrow(SR.dat)

nrow(VWC.dat)
nrow(VWC.dat.new)
nrow(VWC.dat.new)/nrow(VWC.dat)

# check how many pts per del period
# dat<-GCC.dat.new
# dat<-SR.dat.new
dat<-VWC.dat.new

point.counts<-dat%>%
  dplyr::count(Study_shorthand, year, trt, resp_period)%>% # how many data points are in eacy period
  arrange(resp_period,Study_shorthand)
point.counts

# check any studies missing boundaries
dat%>%
  group_by(Study_shorthand,year,trt)%>%
  summarise(
    del_start=first(del_start),
    resp_period_end=first(resp_period_end),
    first_date=min(date),
    last_date=max(date),
    n=n())%>%
  mutate(missing_boundary=first_date>del_start | last_date<resp_period_end) %>%
  filter(missing_boundary=="TRUE")

#### merge all continuous data together ####
str(VWC.dat.new)
str(GCC.dat.new)
str(SR.dat.new)

VWC.tomerge<-VWC.dat.new%>%
  select(-c(VWC))

GCC.tomerge<-GCC.dat.new%>%
  select(-c(GCC))

SR.tomerge<-SR.dat.new%>%
  select(-c(efflux))%>%
  mutate(plot=as.character(plot))

all.continuous<-bind_rows(VWC.tomerge,GCC.tomerge,SR.tomerge)%>%
  mutate(date=as.Date(date,format="%Y-%m-%d"),
         date=if_else(as.numeric(format(date,"%Y"))<100,
                      date+years(2000),
                      date))
str(all.continuous)

#### averages and plot checks ####
cont.avs<-all.continuous%>%
  group_by(Study_shorthand,Variable,year,date,trt,con_vs_trt,del_size,timing,del_start,del_end,resp_period_end,resp_period)%>%
  dplyr::summarise(
    n.plots.per.date=n(),
    trt.day.av=mean(var,na.rm=TRUE),
    trt.day.SE=(sd(var,na.rm=TRUE))/sqrt(n.plots.per.date))%>%
  filter(!(Study_shorthand=="Tooley_DRE" & year == "2026"))
# view(cont.avs)

## try plotting to check data
### take averages
dat<-cont.avs%>%
  # subset(Variable=="VWC")%>%
  subset(Variable=="GCC")%>%
  # subset(Variable=="SR")%>%
  group_by(Study_shorthand,year)%>%
  mutate(Study_trt=paste0(Study_shorthand,"_",trt),
         trt_type=factor(con_vs_trt,levels=c("control_trt","trt"),labels=c("CON","DEL")))

### specify response period shading
resp_rects<-dat%>%
  filter(!is.na(del_end),!is.na(resp_period_end))%>%
  distinct(Study_shorthand,year,del_start,resp_period_end)
  
### make the plot
plot<-ggplot(dat,aes(x=date,y=trt.day.av,group=trt,linetype=trt_type,color=Study_trt)) +
  geom_rect(data = resp_rects,aes(xmin = del_start,xmax = resp_period_end,ymin = -Inf,ymax = Inf),
            inherit.aes = FALSE,
            fill = "grey80",
            alpha = 0.3)+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=trt.day.av-trt.day.SE, ymax=trt.day.av+trt.day.SE), width=0.2)+
  scale_x_date(date_labels="%b %d") +
  labs(
    x="Date",
    y="VWC",
    linetype="Treatment")+
  theme_bw()+
  theme(legend.position="right")+
  force_panelsizes(rows = unit(2.25, "in"),
               cols = unit(4.25, "in"))+
  facet_wrap(~Study_shorthand + year,scales="free")

plot
```

### Calc. AUC's for all
```{r PREP AUC's - magnitude}
## load package for trapazoidal AUC calculation 
library(pracma)

#### Approximate for AUC boudaries ####
### get daily averages
daily.avs<-all.continuous%>%
  group_by(Variable,Study_shorthand,year,trt,con_vs_trt,date,del_size,timing,del_start,del_end,resp_period_end,resp_period)%>%
  dplyr::summarise(
    n.plots.per.date=sum(!is.na(var)),
    trt.av=mean(var,na.rm=TRUE),
    trt.sd=sd(var,na.rm=TRUE),
    trt.SE=trt.sd/sqrt(n.plots.per.date))

### check studies missing observations before/after boundaries
missing.bounds<-daily.avs%>%
  group_by(Variable,Study_shorthand,year,trt)%>%
  summarise(
    del_start=first(del_start), # first() just grabs the first row's date since all rows repeat the date
    resp_period_end=first(resp_period_end),
    first_date=min(date),
    last_date=max(date),
    n=n())%>%
  mutate(missing_boundary=
           # observations start after deluge started
           first_date>del_start | # OR
           # observations ended before deluge response period ended
           last_date<resp_period_end) %>%
  filter(missing_boundary=="TRUE")
# view(missing.bounds)
  
#### interpolate missing AUC boundaries ####
#### for calculating response period AUC's
auc.dat<-daily.avs%>%
  # drop Tooley GCC 2023-2024 b/c missing boundaries
  filter(!(Study_shorthand=="Tooley_DRE" & Variable == "GCC" & year %in% c(2023,2024)))%>%
  # drop Tooley SR 2024 b/c missing boundaries
  filter(!(Study_shorthand=="Tooley_DRE" & Variable == "SR" & year == "2024"))%>%
  filter(!is.na(del_start),!is.na(resp_period_end))


#### check that grouping combo only have gives one del_start/resp_period_end
check<-auc.dat%>%
  dplyr::count(Variable,Study_shorthand,year,trt,del_start,resp_period_end)%>%
  filter(n > 1)
check # want to be nothing

#### interpolating and adding points on boundaries
## interpolate
auc.dat.bounds<-auc.dat%>%
  group_by(Variable,Study_shorthand,year,trt,del_size,timing,del_start,del_end,resp_period_end)%>%
  # arrange date to put in order (matters for as.numeric(date))
  arrange(date)%>%
  dplyr::summarise(
    start_point=approx(
      x=as.numeric(date),
      y=trt.av,
      xout=as.numeric(first(del_start))
    )$y,
    end_point=approx(
      x=as.numeric(date),
      y=trt.av,
      xout=as.numeric(first(resp_period_end))
    )$y)

# view(auc.dat.bounds)

## make longer (single pt per date)
auc.dat.bounds.long<-auc.dat.bounds%>%
  pivot_longer(cols=c(start_point,end_point),
               names_to="side",
               values_to="trt.av")%>%
  # assign single date to interpolated value for boundary date
  mutate(date=case_when(
    side=="start_point"~del_start,
    side=="end_point"~resp_period_end))%>%
  select(!side)

# view(auc.dat.bounds.long)

## "anti-join" to ONLY keep the interpolated point (auc.dat.bounds.long) if the auc daily avs (auc.dat) do not have a point there already -- so if someone did measure on an exact boundary day, we don't interpolate
auc.new.bounds<-auc.dat.bounds.long%>%
  anti_join(auc.dat,
            by=c("Variable","Study_shorthand","year","trt","del_size","timing","del_start","del_end","resp_period_end","date"))

# view(auc.new.bounds)
nrow(auc.dat.bounds.long) # 184 total interpolated points
nrow(auc.new.bounds) # 116 left when we drop where someone did measure on those dates

## add to the whole data
nrow(auc.dat) # 2030

auc.dat.filled<-bind_rows(
  auc.dat%>%mutate(interpolated=FALSE),
  auc.new.bounds%>%mutate(interpolated=TRUE))%>% # combine
  group_by(Variable,Study_shorthand,year,trt)%>% # sort nicely again
  arrange(date,.by_group=TRUE)%>% # and put dates in order
  ungroup()

nrow(auc.dat.filled) # 2146 == 2030 + 116
```

```{r CHECK AND CALC AUC's - magnitude}
#### take averages
dat<-auc.dat.filled%>%
  subset(Variable=="VWC")%>%
  # subset(Variable=="GCC")%>%
  # subset(Variable=="SR")%>%
  group_by(Study_shorthand,year)%>%
  mutate(Study_trt=paste0(Study_shorthand,"_",trt),
         trt_type=case_when(
           trt %in% c("CON","DR") ~ "CON",
           TRUE ~ "DEL"))%>%
  filter(Study_shorthand %in% c("Hajek2024","Post2021"))

### specify response period shading
resp_rects<-dat%>%
  filter(!is.na(del_end),!is.na(resp_period_end))%>%
  distinct(Study_shorthand,year,del_start,resp_period_end)
  
### make the plot
filled.plot<-ggplot(dat,aes(x=date,y=trt.av,group=trt,linetype=trt_type,color=Study_trt)) +
  geom_rect(data = resp_rects,aes(xmin = del_start,xmax = resp_period_end,ymin = -Inf,ymax = Inf),
            inherit.aes = FALSE,
            fill = "grey80",
            alpha = 0.3)+
  geom_line()+
  geom_point(aes(shape=interpolated), size=3)+
  geom_errorbar(aes(ymin=trt.av-trt.SE, ymax=trt.av+trt.SE), width=0.2)+
  scale_x_date(date_labels="%b %d") +
  labs(
    x="Date",
    y="VWC",
    linetype="Treatment")+
  theme_bw()+
  theme(legend.position="right")+
  force_panelsizes(rows = unit(2.25, "in"),
               cols = unit(4.25, "in"))+
  facet_wrap(~Study_shorthand + year,scales="free")

filled.plot

#### calculate auc's ####
auc.calcs<-auc.dat.filled%>%
  # filter to only dates within the response period based on each row's date and del/resp_period dates
  filter(date>=del_start & date<=resp_period_end)%>%
  group_by(Variable,Study_shorthand,year,trt,del_size,timing,del_start,del_end,resp_period_end)%>%
  # make sure dates are in order
  arrange(date,.by_group=TRUE)%>%
  # calculate!
  dplyr::summarise(
    AUC=trapz(as.numeric(date),trt.av)
  )

auc.calcs.paired <- auc.calcs %>%
  left_join(trt.pairs %>% select(Study_shorthand, year, trt, control_trt, del_start, resp_period_end),
            by = c("Study_shorthand", "year", "trt", "del_start", "resp_period_end"))

# separate trts and controls
auc.trts <- auc.calcs.paired %>%
  filter(!is.na(control_trt))  # only deluge treatments have a control_trt

auc.cons <- auc.calcs.paired %>%
  filter(is.na(control_trt)) %>%
  ungroup() %>%
  select(Variable, Study_shorthand, year, trt, del_start, resp_period_end, AUC) %>%
  rename(AUC_con = AUC, control_trt = trt) %>%
  distinct()

# join by matching control_trt to the control's trt name
auc.diff <- auc.trts %>%
  ungroup() %>%
  left_join(auc.cons,
            by = c("Variable", "Study_shorthand", "year", "control_trt", "del_start", "resp_period_end")) %>%
  mutate(
    AUC_diff     = AUC - AUC_con,
    AUC_day      = AUC / 28,
    AUC_con_day  = AUC_con / 28,
    AUC_diff_day = AUC_diff / 28
  )

str(auc.diff)

auc.diff.clean<-auc.diff%>%
  rename(AUC_del=AUC,AUC_del_div28=AUC_day,AUC_con_div28=AUC_con_day,diff_AUC=AUC_diff,diff_AUC_div28=AUC_diff_day)%>%
  select(Variable,Study_shorthand,year,trt,del_size,timing,del_start,del_end,resp_period_end,AUC_del,AUC_con,diff_AUC)

auc.diff.clean

# write.csv(auc.diff.clean, file = file.path("deluge", "data", "R_exports", "Deluge_response_magnitudes.csv"), row.names = FALSE)
# 
# googledrive::drive_upload(media = file.path("deluge", "data", "R_exports", "Deluge_response_magnitudes.csv"), overwrite = T,
#                         path = googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3"))


  
###### ***export this table***** ######

## check 
# ggplot(auc.diff, aes(x = Study_shorthand, y = AUC_diff_day, color = factor(del_size), shape = timing)) +
#   geom_point(size = 3) +
#   geom_hline(yintercept = 0, linetype = "dashed") +
#   facet_wrap(~Variable, scales = "free_y") +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
# 
# ggplot(auc.diff, aes(x = del_size, y = AUC_diff_day, color = Study_shorthand, shape = timing)) +
#   geom_point(size = 3) +
#   geom_hline(yintercept = 0, linetype = "dashed") +
#   facet_wrap(~Variable, scales = "free_y") +
#   theme_bw()
# 
# r2.labels <- auc.diff %>%
#   group_by(Variable) %>%
#   summarise(
#     r2 = summary(lm(AUC_diff_day ~ del_size))$r.squared,
#     p  = summary(lm(AUC_diff_day ~ del_size))$coefficients[2,4],
#     label = paste0("R² = ", round(r2, 2), ", p = ", round(p, 3))
#   )
# 
# ggplot(auc.diff, aes(x = del_size, y = AUC_diff_day)) +
#   geom_point(aes(color = Study_shorthand, shape = timing), size = 3) +
#   geom_smooth(method = "lm", se = TRUE, aes(group = 1), color = "black", linetype = "dashed") +
#   geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
#   geom_text(data = r2.labels, aes(label = label), x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5) +
#   facet_wrap(~Variable, scales = "free_y") +
#   theme_bw()


##### check response magnitudes vs. size #####
resp_magnitude<-auc.diff%>%
  # filter(timing=="fall")%>%
  filter(!timing=="fall")%>%
  mutate(Variable=factor(Variable,levels=c("VWC","GCC","SR")))%>%
  mutate(Timing=factor(timing,levels=c("early","mid","late","fall"),labels=c("early g.s.","mid g.s.","late g.s.","fall")))

pal<-brewer.pal(3, "Dark2")  # try Dark2, Set2, PRGn, BrBG

library(ggpmisc)

magnitude.plots<-ggplot(dat=resp_magnitude,aes(del_size,AUC_diff_day,shape=Timing,color=del_size))+
  geom_point()+
  scale_shape(name="Timing")+
  scale_x_continuous(limits=c(15,125),n.breaks=7)+
  scale_color_gradient2(low = pal[1], mid = pal[2], high = pal[3], 
                        midpoint = 60,
                        name = "Deluge size (mm)")+
  xlab(expression(bold("Deluge size (mm)")))+
  ylab(expression(atop(bold("Average Daily Deluge Response over 28-day response period"))))+
  geom_smooth(method="lm",color="black",se=FALSE,linewidth=0.7,linetype="dashed",aes(group=1))+
  # ggpmisc::stat_poly_eq(aes(label=paste(after_stat(p.value.label), 
  #                                       after_stat(adj.rr.label),
  #                                       after_stat(rr.label),
  #                                       sep = "~~")),
  #                       size=3,label.x="left",label.y="top")+
  facet_wrap(~Variable,nrow=3,scales="free_y")+
  # geom_vline(xintercept=60)+
  theme_bw()
magnitude.plots

##### export timing vs. relative responses ####
# ggsave()
```



```{r rel 20% - resp lengths}
#### grouping study locations ####
# get max/min for each variable -- range (denominator of DEL-CON/range)
location.range<-all.continuous%>%
  mutate(location=case_when(
    Study_shorthand%in%c("Post2020","Post2021","Condon_SL","Hajek2024","Hoover2022","Tooley_DRE") ~ "IDE",
    Study_shorthand=="CCE" ~ "CCE",
    Study_shorthand=="Linabury_ChANGE" ~ "EDGE"))%>%
  # redo averages (without duplicating controls like daily.avs does)
  group_by(Variable,location,year,date,trt)%>%
  dplyr::summarise(
    n.plots=n(),
    trt.av=mean(var,na.rm=TRUE))%>%
  # use non-duplicated control av's to get max and min
  group_by(Variable,location,year)%>%
  dplyr::summarise(
    min.val=min(trt.av, na.rm=TRUE),
    max.val=max(trt.av, na.rm=TRUE),
    var.range=max.val-min.val)

# update avs too w/ locations -- IDE = Post, Condon, Hajek, Hoover, and Tooley, CCE = Alex only, EDGE = Mary only (these are duplicated avs with extra controls when needed)
daily.avs<-daily.avs%>%
  mutate(location=case_when(
    Study_shorthand %in% c("Post2020","Post2021","Condon_SL","Hajek2024","Hoover2022","Tooley_DRE") ~ "IDE",
    Study_shorthand %in% c("CCE") ~ "CCE",
    Study_shorthand %in% c("Linabury_ChANGE") ~ "EDGE"))


# get deluge and controls side-by-side -- use daily.avs b/c already duplicated controls if needed for multiple del trts
### get correct con for each trt into del (only) data
del.avs<-daily.avs%>%
  filter(con_vs_trt=="trt")%>%
  left_join(trt.pairs%>%
              select(Study_shorthand,year,trt,control_trt,del_start,resp_period_end),
            by=c("Study_shorthand","year","trt","del_start","resp_period_end"))
del.avs

### split out control plots from daily avs too
con.avs<-daily.avs%>%
  ungroup()%>%
  filter(con_vs_trt=="control_trt")%>%
  select(Variable,Study_shorthand,year,trt,date,trt.av) %>%
  rename(con.av=trt.av,control_trt=trt)%>%
  distinct()
con.avs

### join them back together side-by-side correctly using the correct controls for each del trt
wide.daily.avs<-del.avs%>%
  ungroup()%>%
  # combine the controls with the del's data (side-by-side)
  left_join(con.avs,
            by=c("Variable","Study_shorthand","year","control_trt","date"))%>%
  # combine in the correct range for the study's location
  left_join(location.range%>%
              select(Variable,location,year,var.range),
            by=c("Variable","location","year"))%>%
  # calculate the relative difference -- like %difference but divided by range instead of CON again
  mutate(rel.diff=(trt.av-con.av)/var.range)%>%
  mutate(sig.response=case_when(
    (rel.diff*100) >= 30 ~ "pos.30+",
    (rel.diff*100) >= 20 ~ "pos.20+",
    (rel.diff*100) >= 10 ~ "pos.10+",
    (rel.diff*100) <= -30 ~ "neg.30+",
    (rel.diff*100) <= -20 ~ "neg.20+",
    (rel.diff*100) <= -10 ~ "neg.10+",
    TRUE ~ NA))

# view(wide.daily.avs)

## save the dates that are over thresholds
flagged.dates<-wide.daily.avs%>%
  select(Variable,Study_shorthand,year,trt,date,sig.response)%>%
  filter(!is.na(sig.response))

#### plot check relative 20%'s ####
## add flagged dates back into daily.avs to check graphing
dat<-daily.avs%>%
  # subset(Variable=="VWC")%>%
  # subset(Variable=="GCC")%>%
  subset(Variable=="SR")%>%
  group_by(Study_shorthand,year)%>%
  mutate(Study_trt=paste0(Study_shorthand,"_",trt),
         trt_type=case_when(
           trt %in% c("CON","DR") ~ "CON",
           TRUE ~ "DEL"))
nrow(dat)
dat<-dat%>%
  left_join(flagged.dates,by=c("Variable","Study_shorthand","year","trt","date"))
nrow(dat)

### specify response period shading
resp_rects<-dat%>%
  filter(!is.na(del_end),!is.na(resp_period_end))%>%
  distinct(Study_shorthand,year,del_start,resp_period_end)
  
### make the plot
library(ggnewscale)

filled.plot<-ggplot(dat,aes(x=date,y=trt.av,group=trt,linetype=trt_type,color=Study_trt)) +
  geom_rect(data=resp_rects,aes(xmin=del_start,xmax=resp_period_end,ymin=-Inf,ymax=Inf),
            inherit.aes=FALSE,
            fill="grey80",
            alpha=0.3)+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=trt.av-trt.SE, ymax=trt.av+trt.SE), width=0.2)+
  ggnewscale::new_scale_color()+
  geom_text(data=dat%>%filter(!is.na(sig.response)),
            aes(label="+",color=sig.response),size=5,vjust=-0.5)+
  scale_color_manual(values=c(
    "pos.10+" = "darkseagreen2",
    "pos.20+" = "darkseagreen",
    "pos.30+" = "darkseagreen4",
    "neg.10+" = "indianred2",
    "neg.20+" = "indianred3",
    "neg.30+" = "indianred4"
  ))+
  scale_x_date(date_labels="%b %d") +
  labs(
    x="Date",
    y="var",
    linetype="Treatment")+
  theme_bw()+
  theme(legend.position="right")+
  force_panelsizes(rows = unit(2.25, "in"),
               cols = unit(4.25, "in"))+
  facet_wrap(~Study_shorthand + year,scales="free")

filled.plot
```

```{r consecutive 20+ dates}
#### assign response directions ####
response.dat<-wide.daily.avs%>%
  filter(con_vs_trt=="trt")%>%
  mutate(response=case_when(
    rel.diff > 0.20 ~ "positive",
    rel.diff < -0.20 ~ "negative",
    TRUE ~ "none"))%>%
  arrange(Variable,Study_shorthand,year,trt,date)

##### check response lengths #####
gr.dat<-response.dat%>%
  # filter(Variable=="GCC")
  filter(Variable=="VWC")
  # filter(Variable=="SR")

# ggplot(dat=gr.dat,aes(date,rel.diff,color=response))+
#   geom_point()+
#   geom_hline(yintercept=c(-0.2,0.2), linetype="dashed")+
#   facet_wrap(~Study_shorthand+year,scales="free_x")+
#   theme_bw()

resp_period.only<-response.dat%>%
  # filter(resp_period=="del_resp")%>%
  # filter(timing=="fall")%>%
  filter(!timing=="fall")%>%
  mutate(days_since=as.numeric(date)-as.numeric(del_start))%>%
  mutate(Variable=factor(Variable,levels=c("VWC","GCC","SR")))

# y.lims<-c(1,100)
y.lims<-c(1,40)

pal<-brewer.pal(3, "Dark2")  # try Dark2, Set2, PRGn, BrBG

timing.plots<-ggplot(dat=resp_period.only,aes(days_since,rel.diff,shape=control_trt,color=del_size))+
  annotate("rect", xmin=-Inf, xmax=Inf, ymin=-0.2, ymax=0.2, fill="gray80", alpha=0.5)+
  geom_point()+
  scale_shape(name="Background Climate",labels=c("Ambient/LTA","During Drought"))+
  scale_x_continuous(limits=y.lims,n.breaks=7)+
  scale_color_gradient2(low = pal[1], mid = pal[2], high = pal[3], 
                        midpoint = 60,
                        name = "Deluge size (mm)")+
  xlab(expression(bold("Days Since Deluge")))+
  ylab(expression(atop(bold("Increased Deluge Response"), "(Range-normalized % response)")))+
  # facet_wrap(~Variable,nrow=1)+
  geom_vline(xintercept=28)+
  theme_bw()
timing.plots

##### export timing vs. relative responses ####
# ggsave()

# ##### also sort by Study for Q's #####
# resp_period.only<-response.dat%>%
#   # filter(resp_period=="del_resp")%>%
#   # filter(timing=="fall")%>%
#   filter(!timing=="fall")%>%
#   mutate(Study=factor(Study_shorthand,labels=c("Siggers et al. CCE"="CCE","Condon & Knapp 2026"="Condon_SL","Hoover et al. 2024"="Hoover2022","Post & Knapp 2020"="Post2020","Post & Knapp 2021"="Post2021")))%>%
#   # mutate(Study=factor(Study_shorthand,labels=c("Hajek & Knapp 2024"="Hajek2024","Tooley et al. DRE"="Tooley_DRE")))%>%
#   mutate(Variable=factor(Variable,levels=c("VWC","GCC","SR")))%>%
#   mutate(days_since=as.numeric(date)-as.numeric(del_start))
# 
# 
# # y.lims<-c(1,100)
# y.lims<-c(1,40)
# 
# by.study<-ggplot(dat=resp_period.only,aes(days_since,rel.diff,shape=Study,color=trt))+
#   annotate("rect", xmin=-Inf, xmax=Inf, ymin=-0.2, ymax=0.2, fill="gray80", alpha=0.5)+
#   geom_point()+
#   # scale_shape(name="Background Climate",labels=c("Ambient/LTA","During Drought"))+
#   scale_x_continuous(limits=y.lims,n.breaks=7)+
#   xlab(expression(bold("Days Since Deluge")))+
#   ylab(expression(atop(bold("Increased Deluge Response"), "(Range-normalized % response)")))+
#   facet_wrap(~Variable,nrow=1)+
#   # geom_vline(xintercept=28)+
#   theme_bw()
# by.study
# 
# # ggsave()
```

```{r calc response lengths}
#### get consecutive sig.response rates ####
glimpse(response.dat) # this is in start of previous chunk

#### get the previous and next responses
lead.lags.responses<-response.dat%>%
  filter(resp_period %in% c("del_resp","post_del"))%>%
  arrange(Variable,Study_shorthand,year,trt,date)%>%
  group_by(Variable,Study_shorthand,year,trt)%>%
  mutate(prev_resp=lag(response),
         next_resp=lead(response),
         response.filled=case_when(
           ## both sides are none, replace with none
           response !="none" & prev_resp=="none" & next_resp=="none" ~ "none",
           ## both sides are positive, replace with positive
           response !="positive" & prev_resp=="positive" & next_resp=="positive" ~ "positive",
           ## both sides are negative, replace with negative
           response !="negative" & prev_resp=="negative" & next_resp=="negative" ~ "negative",
           ## keep all others same
           TRUE ~ response))

# view(lead.lags.responses)

#### identify consecutive response sequences
library(data.table)

consecutive.responses<-lead.lags.responses%>%
  mutate(run_id=rleid(response.filled))%>%
  group_by(Variable,Study_shorthand,year,trt,run_id)%>%
  mutate(run_response=first(response.filled),
         run_length=n())%>%
  ungroup()

# view(consecutive.responses)

#### summary of run info
response.runs.summary<-consecutive.responses%>%
  
  # drop any runs of "none"
  filter(run_response %in% c("positive","negative"))%>%
  group_by(Variable,Study_shorthand,year,trt,run_id,run_response)%>%
  
  dplyr::summarise(
    # get start date of run
    run_start=min(date),
    # get end date of run
    run_end=max(date),
    # check if end dates ever need to be corrected for ending on a filled date (don't think this is possible from rules)
    end_adj=max(date[response==run_response[1]]),
    # count calendar dates (including start/end date both)
    n_cal_days=as.numeric(run_end-run_start)+1,
    # save response direction
    resp_period=first(resp_period),
    .groups="drop")%>%
  
  # drop any single-day responses (edge cases that weren't filled by rules above b/c prev_resp or next_resp is NA)
  filter(n_cal_days>1)%>%
  
  # within each group, number the runs by start date in case there were multiple rounds of interrupted responses (not sure this ever happens but could be possible ecologically like SoilResp slowing on a few cold days then ramping up again when warm)
  group_by(Variable,Study_shorthand,year,trt,run_response)%>%
  arrange(run_start)%>%
  mutate(run_number=row_number())%>%
  ungroup()

# view(response.runs.summary)

##### identify any lags before consecutive responses #####
lag.days<-consecutive.responses%>%
  
  # now we only want to look at the immediate response period (maybe extend this later for fall)
  filter(!resp_period=="pre_del")%>%
  
  group_by(Variable,Study_shorthand,year,trt)%>%
  dplyr::summarise(
    # find first date of response start
    first_response_date=if_else(
      # if there is a response, grab the min date
      any(response.filled %in% c("positive","negative")),
      min(date[response.filled %in% c("positive","negative")]),
      # else
      as.Date(NA)),
    
    # get first measurement date in the del_resp period
    first_measurement_date=min(date),
    
    # get lag
    lag_days = case_when(
      # when there is a response, count how many days were between first measurement to first response
      is.finite(as.numeric(first_response_date)) ~ as.numeric(first_response_date - first_measurement_date),
      # when there is no response, make the lag NA
      TRUE ~ NA_real_),
    .groups="drop")

# view(lag.days)

#### left join in to daily.avs to get all info together and export
full.response.summary<-auc.diff.clean%>%
  left_join(lag.days%>%
              select(Variable,Study_shorthand,year,trt,first_response_date,first_measurement_date,lag_days),
            by=c("Variable","Study_shorthand","year","trt"))%>%
  left_join(response.runs.summary,
            by = c("Variable", "Study_shorthand", "year", "trt"))%>%
  rename(
    new_resp_run.no=run_number,
    new_resp_start=run_start,
    new_resp_end=run_end,
    new_resp_length=n_cal_days,
    post_del_firstobs_date=first_measurement_date,
    new_resp_lag_days=lag_days)%>%
  mutate(diff_AUC_daily=diff_AUC/28)%>%
  select(Variable,Study_shorthand,year,trt,del_size,timing,
         del_start,del_end,resp_period_end,post_del_firstobs_date,
         new_resp_run.no,new_resp_start,new_resp_end,new_resp_length,new_resp_lag_days,
         AUC_del,AUC_con,diff_AUC,diff_AUC_daily)

# view(full.response.summary)

write.csv(full.response.summary, file = file.path("deluge", "data", "R_exports", "combined_response_magnitude_length.csv"), row.names = FALSE)

googledrive::drive_upload(media = file.path("deluge", "data", "R_exports", "combined_response_magnitude_length.csv"), overwrite = T,
                        path = googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3"))

```


```{r period averages}
## average (pre-deluge, post-deluge through the end of year, response period)

```

```{r ANPP vs. GCC and VWC}








```

```{r GCC vs. VWC, VWC vs. SR}
##### SKIP TO HERE W/O RERUNNING ABOVE ####
mag.length.files<-c("Deluge_response_lengths.csv","Deluge_response_magnitudes.csv","ANPP_paired_aggregated.csv")

for (i in mag.length.files){
  
file.name<-drive.files %>% 
  dplyr::filter(name == i) 

# if-else, to throw error if file name is incorrect or error would otherwise stop loop
if(nrow(file.name) == 1){
  googledrive::drive_download(file = file.name$id, 
                              overwrite = T,
                              path = file.path("deluge", 'data', i))
  
} else{
  message("File error - check name or if duplicated: ", i)
  }
}


# full.response.summary<-read.csv("deluge/data/combined_response_magnitude_length.csv")
# 
# del.mags<-read.csv("deluge/data/Deluge_response_magnitudes.csv")
# glimpse(del.mags)
# 
# del.length<-read.csv("deluge/data/Deluge_response_lengths.csv")
# glimpse(del.length)

anpp<-read.csv("deluge/data/ANPP_paired_aggregated.csv")
glimpse(anpp)
```

### AUC ~ timing and size regressions 
```{r magnitude graphing regressions}
anpp.filt<-anpp%>%
  select(c("Study_shorthand","year","trt","func","ctrl_anpp","ctrl_se","del_anpp","del_se","npp_diff","Size","Difference.from.LTA.MAP","pue","wyr_ppt"))%>%
  rename("del_size"="Size")%>%
  mutate(year=if_else(Study_shorthand %in% c("Hajek2024","Tooley_DRE"),year-1,year))%>%
  mutate(Study_shorthand=str_replace(Study_shorthand,"Hoover2022_CPER_DEX","Hoover2022"))
  
nrow(anpp.filt)

cont.responses<-full.response.summary%>%
  filter(!Study_shorthand=="Tooley_DRE")
nrow(cont.responses)

anpp.filt%>%
  distinct(Study_shorthand,year)

cont.responses%>%
  distinct(Study_shorthand,year)

anpp.cont.combined<-merge(anpp.filt,cont.responses,by=c("Study_shorthand","year","trt","del_size"),all=TRUE)%>%
  filter(!is.na(Variable))
nrow(anpp.cont.combined)

# view(anpp.cont.combined)

#### graph variable magnitude vs. ANPP ####
r2.labels<-anpp.cont.combined%>%
  group_by(Variable,func)%>%
  summarise(
    r2 = summary(lm(npp_diff ~ diff_AUC_daily))$r.squared,
    p  = summary(lm(npp_diff ~ diff_AUC_daily))$coefficients[2,4],
    label = paste0("R² = ", round(r2, 2), ", p = ", round(p, 3)))

gr.dat<-anpp.cont.combined#%>%
  # filter(func=="ANPP")
  # filter(func=="BNPP")
  # filter(func=="FORB")
  # filter(func=="GRASS")

r2.dat<-r2.labels#%>%
  # filter(func=="ANPP")
  # filter(func=="BNPP")
  # filter(func=="FORB")
  # filter(func=="GRASS")

pal<-brewer.pal(3, "Dark2")  # try Dark2, Set2, PRGn, BrBG


ggplot(gr.dat,aes(x = diff_AUC_daily, y = npp_diff)) +
  geom_point(aes(color = del_size, shape = timing), size = 3) +
    scale_color_gradient2(low = pal[1], mid = pal[2], high = pal[3], 
                        midpoint = 60,
                        name = "Deluge size (mm)")+
  geom_smooth(method = "lm", se = FALSE, aes(group = 1), color = "black", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_text(data = r2.dat, aes(label = label), x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5) +
  facet_wrap(~Variable+func, scales = "free") +
  ylab("ANPP Deluge-Control")+
  xlab("Average Daily Magnitude Difference")+
  theme_bw()

#### graph variable magnitude vs. ANPP -- NO CCE ####
r2.labels<-anpp.cont.combined%>%
  # filter(!Study_shorthand=="CCE")%>%
  group_by(Variable,func)%>%
  summarise(
    r2 = summary(lm(npp_diff ~ diff_AUC_daily))$r.squared,
    p  = summary(lm(npp_diff ~ diff_AUC_daily))$coefficients[2,4],
    label = paste0("R² = ", round(r2, 2), ", p = ", round(p, 3)))%>%
  filter(p<0.05)

gr.dat<-anpp.cont.combined%>%
  # filter(!Study_shorthand=="CCE")%>%
  filter(func=="ANPP")
  # filter(func=="BNPP")
  # filter(func=="FORB")
  # filter(func=="GRASS")

r2.dat<-r2.labels%>%
  filter(func=="ANPP")
  # filter(func=="BNPP")
  # filter(func=="FORB")
  # filter(func=="GRASS")

pal<-brewer.pal(3, "Dark2")  # try Dark2, Set2, PRGn, BrBG

ggplot(gr.dat,aes(x=diff_AUC_daily, y=npp_diff)) +
  geom_point(aes(color=del_size, shape=timing), size=3) +
  scale_color_gradient2(low = pal[1], mid = pal[2], high = pal[3], 
                        midpoint = 60,
                        name = "Deluge size (mm)")+
  geom_smooth(data=filter(gr.dat,Variable %in% c("GCC","SR")),method = "lm", se = FALSE, aes(group = 1), color = "black", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_text(data = r2.dat, aes(label = label), x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5) +
  facet_wrap(~Variable, scales = "free")+
  ylab(expression(bold("ANPP Deluge-Control")))+
  xlab(expression(bold("Average Daily Deluge Response over 28 day response period")))+
  theme_bw()

#### **export magnitude vs. anpp** ####



##### check response magnitudes vs. size #####
resp_magnitude<-anpp.cont.combined%>%
  # filter(timing=="fall")%>%
  filter(!timing=="fall")%>%
  mutate(Variable=factor(Variable,levels=c("VWC","GCC","SR")))%>%
  mutate(Timing=factor(timing,levels=c("early","mid","late","fall"),labels=c("early g.s.","mid g.s.","late g.s.","fall")))

pal<-brewer.pal(3, "Dark2")  # try Dark2, Set2, PRGn, BrBG

library(ggpmisc)

magnitude.plots<-ggplot(dat=resp_magnitude,aes(del_size,diff_AUC_daily,shape=Timing,color=del_size))+
  geom_point()+
  scale_x_continuous(limits=c(15,125),n.breaks=7)+
  scale_color_gradient2(low = pal[1], mid = pal[2], high = pal[3], 
                        midpoint = 60,
                        name = "Deluge size (mm)")+
  xlab(expression(bold("Deluge size (mm)")))+
  ylab(expression(atop(bold("Average Daily Deluge Response over 28-day response period"))))+
  geom_smooth(method="lm",color="black",se=FALSE,linewidth=0.7,linetype="dashed",aes(group=1))+
  # ggpmisc::stat_poly_eq(aes(label=paste(after_stat(p.value.label), 
  #                                       after_stat(adj.rr.label),
  #                                       after_stat(rr.label),
  #                                       sep = "~~")),
  #                       size=3,label.x="left",label.y="top")+
  facet_wrap(~Variable,nrow=3,scales="free_y")+
  # geom_vline(xintercept=60)+
  theme_bw()
magnitude.plots

##### **export timing vs. relative responses** ####
# ggsave()

```

```{r timing graphing regressions}
r2.labels<-anpp.cont.combined%>%
  group_by(Variable,func)%>%
  summarise(
    r2 = summary(lm(npp_diff ~ diff_AUC_daily))$r.squared,
    p  = summary(lm(npp_diff ~ diff_AUC_daily))$coefficients[2,4],
    label = paste0("R² = ", round(r2, 2), ", p = ", round(p, 3)))

gr.dat<-anpp.cont.combined#%>%
  # filter(func=="ANPP")
  # filter(func=="BNPP")
  # filter(func=="FORB")
  # filter(func=="GRASS")

r2.dat<-r2.labels#%>%
  # filter(func=="ANPP")
  # filter(func=="BNPP")
  # filter(func=="FORB")
  # filter(func=="GRASS")

######## NOTES: add correct color scale from earlier plots here & run again w/o CCE #########
pal<-brewer.pal(3, "Dark2")  # try Dark2, Set2, PRGn, BrBG


ggplot(gr.dat,aes(x = new_resp_length, y = npp_diff)) +
  geom_point(aes(color = del_size, shape = timing), size = 3) +
  scale_color_gradient2(low = pal[1], mid = pal[2], high = pal[3], 
                        midpoint = 60,
                        name = "Deluge size (mm)")+
  geom_smooth(method = "lm", se = FALSE, aes(group = 1), color = "black", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_text(data = r2.dat, aes(label = label), x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5) +
  facet_wrap(~Variable+func, scales = "free_y") +
  ylab(expression(bold("Productivity Difference (g m"^-2*")")))+
  xlab("Average Daily Deluge Response over 28-day response period")+
  theme_bw()

```


### VPD maybe? vs. VWC?? vs. GCC?? vs. ANPP??
How much does VWC increase depends on VPD

