---
title: "VWC Extraction"
author: "OLH"
date: "2026-02-2020"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, supportR, lubridate)

`%!in%` = Negate(`%in%`)

```

# Create Folders on Desktop
```{r}
dir.create(file.path("deluge"), showWarnings = F)
dir.create(file.path("deluge", 'data'), showWarnings = F)
dir.create(file.path("deluge", 'data', 'R_exports'), showWarnings = F)
```

# Download VWC files
```{r}
# Identify wanted files
files_drive <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1-z0EMfF8_LTWr3HZVajzcO3K9jKgYM9p")) %>% 
  dplyr::filter(stringr::str_detect(string = .$name, pattern = "VWC.*\\.csv|\\.csv.*VWC"))

# Did that work?
files_drive

# Identify local files
files_local <- dir(path = file.path("deluge", "data"))
files_local

# Overwrite local data files?
update <- TRUE

# Identify desired files
if(update == T) {
  files_wanted <- files_drive 
} else {
  files_wanted <- files_drive %>%
    dplyr::filter(!name %in% files_local)
}

# Download them!
purrr::walk2(.x = files_wanted$id, .y = files_wanted$name,
             .f = ~ googledrive::drive_download(file = .x, overwrite = T,
                                                path = file.path("deluge", "data", .y)))


```

# Read in all of the data
```{r}

# Get list of files with VWC in the filename
vwc_files <- list.files(path = file.path("deluge", "data"), 
                        pattern = "VWC", 
                        full.names = TRUE)

# Read each file and assign to a dataframe named after the file
for(f in vwc_files) {
  # Extract just the filename without extension as the object name
  obj_name <- tools::file_path_sans_ext(basename(f))
  
  # Read the CSV and assign to an object with that name
  assign(obj_name, read.csv(f))
}

```

# Data processing each site
- Want to get where it is site_id, var, treatment, date, year, doy, value at the plot

## Hajek2024
```{r}
hajek2024 <- Hajek2024_VWC %>%
  rename(plot = "Plot", 
         trt = "Trt", 
         date = "Date") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Hajek2024", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%d-%b-%y"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA,
         VWC=as.numeric(VWC)
         )

# double-check rows where as.numeric conversion failed
Hajek2024_VWC %>%
  mutate(var = suppressWarnings(as.numeric(VWC))) %>%
  filter(is.na(var) & !is.na(VWC))  # NA in result but NOT NA in original = coercion failure - all look OK for Hajek n/a --> NA
  
str(hajek2024)

# check averages
hajek.avs<-hajek2024%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=sum(!is.na(VWC)),
    VWC.av=mean(VWC,na.rm=TRUE),
    VWC.sd=sd(VWC,na.rm=TRUE),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))

# view(hajek.avs) # 9/17/21 -- drop b/c during deluge addition

hajek2024<-hajek2024%>%
  filter(!date=="2021-09-17")

str(hajek2024) # dropped 9/17/21
```

## Hoover et al., 2022
- this is actually sentek; need to go back and see if can do non-sentek stuff
```{r}
str(Hoover2022_VWC_20cm)

hoover.long <- Hoover2022_VWC_20cm %>%
 pivot_longer(
    cols = starts_with("X"),
    names_to = "Date",
    values_to = "VWC"
  ) %>%
  mutate(
    Date = stringr::str_remove(Date, "^X"),
    Date = stringr::str_replace_all(Date, "\\.", "-"),
    # Add "-19" if no year supplied
    Date = ifelse(!grepl("\\d{2}$", Date), paste0(Date, "-19"), Date),
    Date = as.Date(Date, format = "%d-%b-%y")
  ) %>%
  # remove na
  filter(!is.na(VWC))

str(hoover.long)

hoover2022 <- hoover.long %>%
  rename(trt = "TRT",
         plot = "Plot",
         date = "Date") %>%
    # add in the shorthand name
  mutate(Study_shorthand = "Hoover2022",
         # add in the variable analyzed
         Variable = "VWC",
         year = year(date),
         doy = yday(date),
         block = NA, 
         trt = as.character(trt),
         trt = case_when(
         trt == 1 ~ "AMB",
         trt == 2 ~ "CON", 
         trt == 3 ~ "DRT", 
         trt == 4 ~ "DRT+DEL", 
         trt == 5 ~ "DRT+SE", 
         TRUE ~ trt   
         ), 
         study_trt = paste(Study_shorthand, trt, sep="_"))

str(hoover2022)

# check for NA's
hoover2022%>%
  filter(is.na(VWC)) # none, dropped above
  
# check averages
hoover.avs<-hoover2022%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=n(),
    VWC.av=mean(VWC),
    VWC.sd=sd(VWC),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))

# view(hoover.avs) # only a couple of days with half plots (n=5/trt instead of 10) -- okay and leaving
  
```
## Post2020
```{r}
str(Post2020_VWC)

post2020 <- Post2020_VWC %>%
  rename(plot = "Plot", 
         trt = "Trt", 
         date = "Date") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Post2020", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%Y"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA)
str(post2020)

# check for NA's
post2020%>%
  filter(is.na(VWC)) # none

# check averages
post20.avs<-post2020%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=n(),
    VWC.av=mean(VWC),
    VWC.sd=sd(VWC),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))

# view(post20.avs) # all dates have 6-8 plots, okay and leaving
```

## Post 2021

```{r}
str(Post2021_VWC)

post2021 <- Post2021_VWC %>%
  rename(plot = "Plot", 
         trt = "Trt", 
         date = "Date") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Post2021", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = dmy(date), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA)
str(post2021)

# check NA's
post2021%>%
  filter(is.na(VWC)) # none

# check averages
post21.avs<-post2021%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=n(),
    VWC.av=mean(VWC),
    VWC.sd=sd(VWC),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))

# view(post21.avs) # always 4, OK and leaving
```

## CCE - Siggers
```{r}
str(CCE_VWC_Year2)

cce <- CCE_VWC_Year2 %>%
  rename(plot = "Plot", 
         trt = "Treatment", 
         date = "Date", 
         VWC = "X20_cm_VWC.", 
         block = "Block") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "CCE", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%Y-%m-%d"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_")) %>%
  select(-Notes)
str(cce)

# check na's
cce%>%
  filter(is.na(VWC)) # 71 NA rows -- but all at least n=5/trt from avs below

view(cce)
# check averages
cce.avs<-cce%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=sum(!is.na(VWC)),
    VWC.av=mean(VWC,na.rm=TRUE),
    VWC.sd=sd(VWC,na.rm=TRUE),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))

# view(cce.avs) - all dates look OK - at least n=5/trt -- leaving alone
```

## Condon

```{r}
str(Condon_SLs_VWC)

condon_sl <- Condon_SLs_VWC %>%
  rename(plot = "Plot", 
         trt = "NewTrt", 
         date = "Date") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Condon_SL", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%y"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA) %>%
  select(-c(X, Comb.TRT))

str(condon_sl)

# check na's
condon_sl%>%
  filter(is.na(VWC)) # 12 lines of NA's -- looks like one day w/ lots of plots missed but on av's there are still 8-9 plots/trt

# check avs
condon.avs<-condon_sl%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=sum(!is.na(VWC)),
    VWC.av=mean(VWC,na.rm=TRUE),
    VWC.sd=sd(VWC,na.rm=TRUE),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))

# view(condon.avs) # all looks OK -- leaving alone

```


## Linabury
```{r}
str(Linabury_ChANGE_VWC)

linabury <- Linabury_ChANGE_VWC %>%
  rename( 
         trt = "treatment", 
         VWC = "vwc") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Linabury_ChANGE", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%Y"), 
         study_trt = paste(Study_shorthand, trt, sep="_")) %>%
  select(-c(X, site, deluge.date.1, deluge.date.2, dom, month, days_since_deluge, experiment, nitrogen))

str(linabury)

# check na's 
linabury%>%
  filter(is.na(VWC)) # none

# check avs
lin.avs<-linabury%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=n(),
    VWC.av=mean(VWC),
    VWC.sd=sd(VWC),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))

# view(lin.avs) # 7/27, 8/10, and 8/28 only 1 plot per trt measured

# check again using days_since_deluge instead of calendar dates
linabury.2<-Linabury_ChANGE_VWC%>%
  rename( 
         trt = "treatment", 
         VWC = "vwc") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Linabury_ChANGE", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%Y"), 
         study_trt = paste(Study_shorthand, trt, sep="_")) %>%
  select(-c(X, site, deluge.date.1, deluge.date.2, dom, month, experiment, nitrogen))

linabury.avs2<-linabury.2%>%
  group_by(days_since_deluge,trt)%>%
  dplyr::summarise(
    n.plots.per.days=n(),
    VWC.av=mean(VWC),
    VWC.sd=sd(VWC))
# view(linabury.avs2)

# try filtering out the DATES with one plot per and then also check as days_since again
lin.avs<-linabury%>%
  filter(!date %in% c("2021-07-27","2021-08-10","2021-08-28"))%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=n(),
    VWC.av=mean(VWC),
    VWC.sd=sd(VWC),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))
# view(lin.avs)

lin.avs<-linabury.2%>%
  filter(!date %in% c("2021-07-27","2021-08-10","2021-08-28"))%>%
  group_by(days_since_deluge,trt)%>%
  dplyr::summarise(
    n.plots.per.day=n(),
    VWC.av=mean(VWC),
    VWC.sd=sd(VWC),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))
# view(lin.avs) 

## either way w/ filtering, there are either multiple CALENDAR dates with now only n=2/trt or 3, OR multiple DAYS_SINCE_DELUGE dates that have only one plot -- emailed 7/17/2026 to check w/ Mary on data

```

## TOOLEy
```{r}
str(Tooley_DRE_VWC)
tooley <- Tooley_DRE_VWC %>%
  rename(treat=WaterTreatment,legacy=DroughtTreatment)%>%
  mutate(trt2=paste("Leg.",legacy,"_Trt.",treat,sep=""))%>%
  rename(plot = "Plot", 
         trt = "trt2", 
         date = "Date", 
         VWC = "AVGSWC") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Tooley_DRE", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%y"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA) %>%
  select(-c(SWC1, SWC2, SWC3, SWC4, SWC5, SWC6, Notes, ...11, treat, legacy))
str(tooley)

# check na's
tooley%>%
  filter(is.na(VWC)) # lots of NA rows

# check avs
tooley.avs<-tooley%>%
  group_by(date,trt)%>%
  dplyr::summarise(
    n.plots.per.day=sum(!is.na(VWC)),
    VWC.av=mean(VWC,na.rm=TRUE),
    VWC.sd=sd(VWC,na.rm=TRUE),
    VWC.SE=VWC.sd/sqrt(n.plots.per.day))

# view(tooley.avs) # only 2 per trt on 5/15/24, 1/17/25, 6/25/25, drop these dates

str(tooley) # 2097
tooley<-tooley%>%
  filter(!date %in% c("2024-05-15","2025-01-17","2025-06-25"))
str(tooley) # 2025 -- dropped dates OK 
```


# Combine the files

```{r}

combo <- rbind(hajek2024, hoover2022, post2020, post2021, cce, condon_sl, linabury, tooley)
str(combo)

combo%>%
  filter(is.na(VWC)) # 99 NA's OK

#### quick vis together ####
ggplot(combo, aes(doy, VWC))+
  geom_point()+
  theme_bw()+
  facet_wrap(~Study_shorthand, scales="free") # check CCE high outlier and Post 2020 low outlier

combo%>%
  filter(Study_shorthand=="CCE")%>%
  filter(VWC>100) # 2025-09-01 plot 6 129 VWC -- drop this plot

combo%>%
  filter(Study_shorthand=="Post2020")%>%
  filter(VWC<5) # 2017-07-03 plot 31 VWC is 4

combo%>%
  filter(Study_shorthand=="Post2020")%>%
  filter(date=="2017-07-03") # 4 does not look reasonable compared to other plots of same trt on this date

#### drop obv mistake measures ####
str(combo) # 7665

export<-combo%>%
  # drop Post 2017 point
  filter(!(Study_shorthand=="Post2020" & date=="2017-07-03" & plot=="31"))%>%
  # drop CCE point
  filter(!(Study_shorthand=="CCE" & date=="2025-09-01" & plot=="6"))

str(export) # 7663, dropped two rows successfully

# Export locally
write.csv(x = export, row.names = F, na = '',
          file = file.path("deluge", "data", "R_exports", "VWC_20cm_harmonized.csv"))

# Upload to Drive
googledrive::drive_upload(media = file.path("deluge", "data", "R_exports", "VWC_20cm_harmonized.csv"), overwrite = T,
                          path = googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3"))


```

