---
title: "VWC Extraction"
author: "OLH"
date: "2026-02-2020"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, supportR, lubridate)

`%!in%` = Negate(`%in%`)

```

# Create Folders on Desktop
```{r}
dir.create(file.path("deluge"), showWarnings = F)
dir.create(file.path("deluge", 'data'), showWarnings = F)
dir.create(file.path("deluge", 'data', 'R_exports'), showWarnings = F)
```

# Download VWC files
```{r}
# Identify wanted files
files_drive <- googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1-z0EMfF8_LTWr3HZVajzcO3K9jKgYM9p")) %>% 
  dplyr::filter(stringr::str_detect(string = .$name, pattern = "VWC.*\\.csv|\\.csv.*VWC"))

# Did that work?
files_drive

# Identify local files
files_local <- dir(path = file.path("deluge", "data"))
files_local

# Overwrite local data files?
update <- TRUE

# Identify desired files
if(update == T) {
  files_wanted <- files_drive 
} else {
  files_wanted <- files_drive %>%
    dplyr::filter(!name %in% files_local)
}

# Download them!
purrr::walk2(.x = files_wanted$id, .y = files_wanted$name,
             .f = ~ googledrive::drive_download(file = .x, overwrite = T,
                                                path = file.path("deluge", "data", .y)))


```

# Read in all of the data
```{r}

# Get list of files with VWC in the filename
vwc_files <- list.files(path = file.path("deluge", "data"), 
                        pattern = "VWC", 
                        full.names = TRUE)

# Read each file and assign to a dataframe named after the file
for(f in vwc_files) {
  # Extract just the filename without extension as the object name
  obj_name <- tools::file_path_sans_ext(basename(f))
  
  # Read the CSV and assign to an object with that name
  assign(obj_name, read.csv(f))
}

```

# Data processing each site
- Want to get where it is site_id, var, treatment, date, year, doy, value at the plot

## Hajek2024
```{r}
hajek2024 <- Hajek2024_VWC %>%
  rename(plot = "Plot", 
         trt = "Trt", 
         date = "Date") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Hajek2024", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%d-%b-%y"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA)
  
str(hajek2024)
```

## Hoover et al., 2022
- this is actually sentek; need to go back and see if can do non-sentek stuff
```{r}
str(Hoover2022_VWC_20cm)

hoover.long <- Hoover2022_VWC_20cm %>%
 pivot_longer(
    cols = starts_with("X"),
    names_to = "Date",
    values_to = "VWC"
  ) %>%
  mutate(
    Date = stringr::str_remove(Date, "^X"),
    Date = stringr::str_replace_all(Date, "\\.", "-"),
    # Add "-19" if no year supplied
    Date = ifelse(!grepl("\\d{2}$", Date), paste0(Date, "-19"), Date),
    Date = as.Date(Date, format = "%d-%b-%y")
  ) %>%
  # remove na
  filter(!is.na(VWC))

str(hoover.long)
hoover2022 <- hoover.long %>%
  rename(trt = "TRT",
         plot = "Plot",
         date = "Date") %>%
    # add in the shorthand name
  mutate(Study_shorthand = "Hoover2022",
         # add in the variable analyzed
         Variable = "VWC",
         year = year(date),
         doy = yday(date),
         block = NA, 
         trt = as.character(trt),
         trt = case_when(
         trt == 1 ~ "AMB",
         trt == 2 ~ "CON", 
         trt == 3 ~ "DRT", 
         trt == 4 ~ "DRT+DEL", 
         trt == 5 ~ "DRT+SE", 
         TRUE ~ trt   
         ), 
         study_trt = paste(Study_shorthand, trt, sep="_"))
str(hoover2022)
```
## Post2020
```{r}
str(Post2020_VWC)

post2020 <- Post2020_VWC %>%
  rename(plot = "Plot", 
         trt = "Trt", 
         date = "Date") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Post2020", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%Y"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA)
str(post2020)
```

## Post 2021

```{r}
str(Post2021_VWC)

post2021 <- Post2021_VWC %>%
  rename(plot = "Plot", 
         trt = "Trt", 
         date = "Date") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Post2021", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = dmy(date), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA)
str(post2021)
```
## CCE - Siggers
```{r}
str(CCE_VWC_Year2)

cce <- CCE_VWC_Year2 %>%
  rename(plot = "Plot_.", 
         trt = "Trt_.", 
         date = "Date", 
         VWC = "X20_cm_VWC_.") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "CCE", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%y"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA) %>%
  select(-X)
str(cce)
```

## Condon

```{r}
str(Condon_SLs_VWC)

condon_sl <- Condon_SLs_VWC %>%
  rename(plot = "Plot", 
         trt = "NewTrt", 
         date = "Date") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Condon_SL", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%y"), 
         year = year(date), 
         doy = yday(date), 
         study_trt = paste(Study_shorthand, trt, sep="_"), 
         block = NA) %>%
  select(-c(X, Comb.TRT))

str(condon_sl)
```


## Linabury
```{r}
str(Linabury_ChANGE_VWC)

linabury <- Linabury_ChANGE_VWC %>%
  rename( 
         trt = "treatment", 
         VWC = "vwc") %>%
  # add in the shorthand name
  mutate(Study_shorthand = "Linabury_ChANGE", 
         # add in the variable analyzed
         Variable = "VWC", 
         date = as.Date(date, format = "%m/%d/%Y"), 
         study_trt = paste(Study_shorthand, trt, sep="_")) %>%
  select(-c(X, site, deluge.date.1, deluge.date.2, dom, month, days_since_deluge, experiment, nitrogen))

str(linabury)
```

# Combine the files

```{r}

combo <- rbind(hajek2024, hoover2022, post2020, post2021, cce, condon_sl, linabury)
str(combo)

# Export locally
write.csv(x = combo, row.names = F, na = '',
          file = file.path("deluge", "data", "R_exports", "VWC_20cm_harmomnized.csv"))

# Upload to Drive
googledrive::drive_upload(media = file.path("deluge", "data", "R_exports", "VWC_20cm_harmomnized.csv"), overwrite = T,
                          path = googledrive::as_id("https://drive.google.com/drive/folders/1dmzGwkGC1Y4uWW8nLXDkhtB_4A4suyF3"))


```


